Identification of the trail-following pheromone receptor in termites

Version 1

Abstract

Abstract

Chemical communication is the cornerstone of eusocial insect societies since it mediates the social hierarchy, division of labor, and concerted activities of colony members. The chemistry of social insect pheromones received considerable attention in both major groups of social insects, the eusocial Hymenoptera and termites. By contrast, current knowledge on molecular mechanisms of social insect pheromone detection by odorant receptors (ORs) is limited to hymenopteran social insects and no OR was yet functionally characterized in termites, the oldest eusocial insect clade. Here, we present the first OR deorphanization in termites. Using the data from antennal transcriptome and genome of the termite Prorhinotermes simplex (Rhinotermitidae), we selected 4 candidate OR sequences, expressed them in Empty Neuron Drosophila, and functionally characterized using single sensillum recording (SSR) and a panel of termite semiochemicals. In one of the selected ORs, PsimOR14, we succeeded in obtaining strong and reliable responses to the main component of P. simplex trail-following pheromone, the monocyclic diterpene neocembrene. PsimOR14 showed a narrow tuning to neocembrene; only one additional compound out of 72 tested (geranylgeraniol) generated non-negligible responses. Subsequently, we used SSR and P. simplex workers to identify the olfactory sensillum specifically responding to neocembrene, thus likely expressing PsimOR14. We report on homology-based modelling of neocembrene binding by PsimOR14 and show how different ligands impact the receptor dynamicity using molecular dynamics simulations. Finally, we demonstrate that PsimOR14 is significantly more expressed in workers than in soldiers, which correlates with higher sensitivity of workers to neocembrene.

Introduction

Chemical communication is the cornerstone of eusocial insect societies. It mediates the social hierarchy and division of labor, social cohesion and concerted activities of colony members. Major clades of eusocial insects have convergently evolved chemical signals acting in similar social contexts, as manifested by the presence of queen pheromones, trail-following pheromones, alarm pheromones, sex pheromones, food-marking pheromones, and colony and/or caste recognition cues in eusocial Hymenoptera and in termites, in spite of different glandular origins and diverse chemistries of these signals in the two evolutionarily distant insect clades (Leonhardt et al., 2016).

Chemical diversity of pheromone communication in termites (Isoptera) has been extensively studied during the past decades. This applies in particular to trail-following pheromones (TFPs) and sex-pairing pheromones (SPPs), whose current knowledge covers all major termite lineages and allows for evolutionary reconstructions of pheromone chemistry. In contrast to the overall complexity of chemical communication in termite colonies, the chemical diversity of TFPs and SPPs is relatively modest, with only a handful of compounds acting as these pheromones or as components thereof across the phylogeny of Isoptera. Moreover, due to common evolutionary origin of trail-following behavior and sex-pairing behavior, the same compounds may be parts of both SPPs and TFPs in individual species and lineages, adding yet another level of pheromone parsimony (Bordereau & Pasteels, 2011). In the basal lineages Mastotermitidae + Teletisoptera, linear and branched primary alcohols and aldehydes with 13 or 14 carbons play the role of TFPs and SPPs. Major evolutionary transition takes place in Kalotermitidae + Neoisoptera (Icoisoptera, over 98% of living species), in which unbranched C12 fatty alcohols with one, two or three double bonds became almost universally present in both pheromone types as single components or in species-specific mixtures (Bagnères & Hanus, 2015; Bordereau & Pasteels, 2011; Mitaka & Akino, 2021; Sillam-Dussès, 2010). Within Neoisoptera, the recently diversified group containing over 70% of modern termite species, this striking conservativeness of TFP and SPP components is in part enriched by occasional occurrence of terpenoids, such as the cyclic diterpene hydrocarbons neocembrene and trinervitatriene (Bordereau & Pasteels, 2011; Mitaka & Akino, 2021).

Since the identification of the multigene family of insect odorant receptors (ORs) expressed in antennae and maxillary palps of Drosophila melanogaster (Clyne et al., 1999), neurophysiology of insect olfaction underwent a great progress manifested by phylogenetic reconstructions of their evolution across insect taxa, functional characterizations (deorphanisations) of many ORs and ultimately by structural characterizations (Butterwick et al., 2018; del Mármol et al., 2021). Even though the insect ORs have seven transmembrane domains (TMDs) like the ORs known from vertebrates, they do not share many tangible sequence similarities. Insect ORs have an inverted membrane topology in the dendrites of olfactory sensory neurons compared to vertebrate ORs (Benton et al., 2006; Clyne et al., 1999) and unlike the vertebrate ORs known to act as GPCR receptors, insect ORs function as odorant-gated ion channels (Sato et al., 2008; Wicher et al., 2008). While the ORs from the basal wingless insect lineage Archaeognatha are organized as homotetramers (del Mármol et al., 2021) and lack a coreceptor protein (Brand et al., 2018), ORs of all other Insecta including the apterygote clade of Zygentoma form heteromeric complexes with a highly conserved coreceptor protein (ORCo) (Brand et al., 2018; Butterwick et al., 2018; Larsson et al., 2004; Sato et al., 2008).

The OR repertoire is greatly variable across Insecta and ranges from units in the wingless Archaeognatha and basal winged order Odonata to tens or hundreds ORs identified in most other flying insects, the record holders being Hymenoptera, especially ants with up to five hundred identified ORs (Engsontia et al., 2015; Robertson, 2019; Yan et al., 2020). Insect ORs often lack a clear orthology pattern across phylogeny and the OR family evolved via rapid birth-and-death process, accompanied by multiple gene duplications, pseudogenizations, and losses. Lineage-specific expansions, together with a considerable variability in OR ligand specificities (from broad to narrow tuning) allow for a rapid response of olfactory system to ecological and life history changes (Andersson et al., 2015; Benton, 2015; Nei & Rooney, 2005; Robertson, 2019). Ecology-driven plasticity in OR evolution has been convincingly demonstrated by comparisons of specialist vs. generalist insects, the former often having much lower repertoires of ORs and other chemosensory proteins such as gustatory (GR) and ionotropic receptors (IR) (Robertson, 2019). Another plausible example of OR evolution responding to life history traits is that of eusocial Hymenoptera; their complex chemical communication linked with eusociality is accompanied by striking OR family expansion (Engsontia et al., 2015; Legan et al., 2021; McKenzie et al., 2016; Pask et al., 2017; Zhou et al., 2015).

The knowledge on OR structure and function, ligand specificities and tuning has historically been obtained mainly from studies on holometabolan insects using various heterologous deorphanisation systems, and eusocial Hymenoptera received considerable attention with multiple ORs being deorphanized in ants (Pask et al., 2017; Slone et al., 2017) and the honey bee (e.g., Gomez Ramirez et al., 2023; Wanner et al., 2007). Only recently this bias was in part compensated by successful OR deorphanisations in more basal taxa, e.g., the basal insect lineage Archaeognatha (del Mármol et al., 2021) or the hemimetabolous aphids (Zhang et al., 2017, 2019). Within Polyneoptera, OR detecting the aggregation pheromone was identified in the migratory locust (Guo et al., 2020) and very recently a comprehensive deorphanization effort identified multiple narrowly tuned locust ORs responsible for environmental cues recognition (Chang et al., 2023). No attempt was yet made to functionally characterize the ORs in termites, the oldest clade of social insects, which evolved roughly 140 my within Polyneoptera as an internal group of cockroaches (Blattodea) (Buček et al., 2019; Evangelista Dominic et al., 2019). Diversity of termite ORs and other chemoreceptor proteins was previously inferred from genome assemblies (Harrison et al., 2018; Terrapon et al., 2014) or whole-body transcriptomes (Mitaka et al., 2016). Recently, we performed a comprehensive search for chemosensory protein repertoire in workers of three termite species belonging to different families using the antennal transcriptomes (Johny et al., 2023). At the first view, the correlation between the social evolution and OR diversity depicted for eusocial Hymenoptera does not hold in termites, because termite ORs are organized in relatively conserved, highly orthologous pattern and their total numbers, ranging from 28 to 69 (Johny et al., 2023; Terrapon et al., 2014), are quite low and even lower than OR repertoire of the solitary cockroach Blattella germanica (Harrison et al., 2018).

Here, we report on the first deorphanisation of an OR from termites, more specifically on the OR for the major component of the TFP in Prorhinotermes simplex (Rhinotermitidae), the cyclic diterpene neocembrene. We take advantage of the good knowledge on the chemical ecology of this species, ranging from the chemistry of soldier-produced defensive compounds (Hanus et al., 2006; Jirošová et al., 2017; Piskorski et al., 2007) to the chemical identity of TFPs and SPPs. The sex-pairing behavior is mediated by the C12 alcohol (3Z,6Z,8E)-dodecatrien-1-ol (Hanus et al., 2009). The same compound also serves as the minor component of TFP, the major component being neocembrene (Sillam-Dussès et al., 2009). We build on the annotated repertoire of 50 ORs from P. simplex antennal transcriptome (Johny et al., 2023) and our additional P. simplex sequencing data, i.e., caste-specific head transcriptomes, draft genome assembly and long-term laboratory culture of the species in multiple colonies. We select four P. simplex candidate OR sequences from reconstructed termite OR phylogeny, study their function by means of the Empty Neuron Drosophila in vivo expression system and single-sensillum recording (SSR) using panels of potential ligands with biological relevance for termites. We report on strong and selective response of PsimOR14 to neocembrene. In addition, we identify the P. simplex olfactory sensillum specifically responding to neocembrene and document worker caste-biased expression of PsimOR14 accompanied by significantly higher sensitivity to neocembrene, compared to P. simplex soldiers.

Results

Phylogenetic reconstruction and candidate OR selection

In the first step, we reconstructed the phylogeny of termite ORs using previously published OR protein sequences from two species in combination with our antennal transcriptome data on workers from three species (Johny et al., 2023), including P. simplex. The resulting phylogenetic tree, shown in Fig. 1 and Supplementary Fig. S1, revealed an expected highly orthologous pattern, with OR sequences from all five termite species being represented in most sub-branches and with relatively few species-specific expansions. Beside two basally situated monophyletic lineages consisting of only a few OR sequences, the remainder of the tree was organized into two large sister clusters rich in ORs from all studied species. From the 50 P. simplex ORs mapped on the tree, we selected for the first deorphanization attempts 4 ORs (PsimOR9, 14, 30 and 31) distributed at different branches across the phylogenetic diversity of termite ORs (Fig. 1A). Their sequences were used for transgenic D. melanogaster generation and SSR screening.

Phylogenetic reconstruction of termite ORs

The tree is based on 182 protein sequences from five species of termites and the bristletail Lepisma saccharina as a basal insect outgroup, and also includes the sequences of OR co-receptor (OrCO). The topology and branching supports were inferred using the IQ-TREE maximum likelihood algorithm with the JTT+F+R8 model and supported by 10,000 iterations of ultrafast bootstrap approximation. Protein sequences of termite ORs can be found under the same labeling in Johny et al. (2023). Lepisma saccharina sequences are listed in Thoma et al. (2019). Arrowheads highlight the four ORs from Prorhinotermes simplex selected for functional characterization. Fully annotated version of the tree is provided as Supplementary Fig. S1. Heatmap shows the abundances of 50 OR transcripts identified in the RNAseq data from P. simplex antennae available in NCBI SRA archive under accession SRX17749141.

Functional characterization of P. simplex ORs in D. melanogaster ab3 sensillum

We expressed the four selected P. simplex ORs in the recently improved version of the Drosophila melanogaster Empty Neuron system (Chahda et al., 2019); the crossing scheme for fly generation adapted from Gonzales et al. (Gonzalez et al., 2016) is shown in Supplementary Fig. S2. Spontaneous SSR firing rates of the four transgenic lines showed an expected pattern with no abnormal bursts, indicating that the ORs were functional. The flies were first subjected to SSR screening with Panel 1, consisting of 17 semiochemicals relevant to termite chemical communication and structurally related compounds. As shown in Fig. 2A, PsimOR9 and PsimOR30 did not provide any strong response to any of the tested compounds, while PsimOR31 broadly and weakly responded to several compounds. By contrast, PsimOR14 systematically and strongly responded to stimulations by the monocyclic diterpene hydrocarbon neocembrene. Additionally, a moderate PsimOR14 response was also recorded for the linear diterpene alcohol geranylgeraniol, while all other compounds in the panel, including three other terpenoids, only elicited weak or no responses (Fig. 2A, Supplementary Tables S1–S4).

SSR responses of transgenic D. melanogaster ab3 sensillum expressing PsimOR9, 14, 30 and 31 to the panel of 17 volatiles with biological relevance for termites

A Heatmap showing the average responses of the four ORs as Δ spikes/s from 3–6 independent replicates. B. Comparison of SSR responses of transgenic Drosophila melanogaster ab3A neurons expressing PsimOR14 (ab3A:PsimOR14) and W1118 D. melanogaster. The bars show the average Δ spikes/s values from five independent replicates ± SEM. C. Characteristic SSR traces of ab3A:PsimOR14 and W1118 flies for 1 µg dose of neocembrene and geranylgeraniol. D. Dose response curve of ab3A:PsimOR14 SSR responses to neocembrene. The graph shows average Δ spikes/s values ± SEM based on nine replicates (8 in case of 100 ng and 4 in case of 500 ng stimulations). The curve fit and ED50 value were calculated using log(agonist) vs. response non-linear algorithm with least square fit method and the constraint of minimal response > 0. The raw data for all graphs is provided in Supplementary Tables S1–S6.

We then compared the responses of Drosophila ab3 sensillum in PsimOR14 expressing flies with those of W1118 flies. As evidenced in Fig. 2B, W1118ab3 sensillum did not show any significant neuronal response to Panel 1 compounds, while the transgenic PsimOR14 line generated an average Δ spike number of >50 spikes/s for neocembrene and a minor secondary response of ∼25 Δ spikes/s for geranylgeraniol. Characteristic responses for both lines to the two compounds are depicted in Fig. 2C. In the next step, we tested the dose-response behavior of PsimOR14 flies to neocembrene, which showed an exponentially increasing neuronal response over the range of 0.01–10 ng to ED50 = ∼24 ng and a lack of saturation at the dose of 500 ng (Fig. 2D, Supplementary Tables S5, S6).

PsimOR14 is narrowly tuned to neocembrene, the main trail-following pheromone component

To further address the specificity of PsimOR14 tuning, we tested three additional panels containing 56 frequently occurring insect semiochemicals from various chemical classes. As shown in Fig. 3A, none of these compounds, including multiple terpenoids (mono-, sesqui-, di-), generated a strong response, suggesting a narrow tuning of PsimOR14 to neocembrene. The narrow tuning of PsimOR14 is evident also from the tuning curve depicted in Fig. 3B, with the receptor lifetime sparseness value 0.89 (Supplementary Tables S7–S9). These results suggest that PsimOR14 is a pheromone receptor adaptively tuned to neocembrene, which is the main component of the trail-following pheromone in the genus Prorhinotermes.

SSR responses of transgenic D. melanogaster ab3 sensillum expressing PsimOR14 to the complete set of 73 compounds (Panels 1–4)

A. Heatmap showing the average responses as Δ spikes/s from 3–6 independent replicates. C. Tuning curve of PsimOR14 for the 73 compounds contained in panels 1–4. The raw data is provided in Supplementary Tables S7–S9.

Identification of P. simplex olfactory sensillum involved in neocembrene detection

Our next goal was to identify the antennal olfactory sensillum responsible for neocembrene detection by P. simplex workers by combination of SSR measurements with SEM and HR-SEM imaging. Since no previous study reported SSR responses of termite olfactory sensilla to pheromones and environmental cues, we decided to search for the neocembrene-detecting sensillum on the last flagellomere, known to harbor by far the most sensilla in termite workers (Castillo et al., 2021).

SEM mapping of the distal antennal segment revealed multiple types of multiporous trichoid and basiconic sensilla with potential olfactory function (Fig. 4A). In SSR experiments using the termite-relevant compounds from Panel 1, we obtained strong response to both neocembrene and geranylgeraniol from a short multiporous grooved sensillum situated in the apical part of the last segment. Detailed retrieval using HR-SEM allowed us to distinguish five types of multiporous grooved sensilla according to their wall structure, overall shape, length, and shape of the sensillar tip. The neocembrene-detecting sensillum belongs to S I type according to this classification (Fig. 4A, B). Detailed view on spontaneous firing pattern of the neocembrene-detecting sensillum revealed three different spike amplitudes (Fig. 4C), suggesting the potential presence of as many as three olfactory sensory neurons (a–c). Comparison with spike amplitudes upon neocembrene and geranylgeraniol stimulation then indicated that the responses correspond to the neuron labeled as b.

Distribution of olfactory sensilla on the last antennal segment of P. simplex workers

A. Arrows show the localization of selected representatives of the five distinguished multiporous grooved sensilla types (S I–V) on a SEM photograph of the last flagellomere. Scale bar represents 50 µm. B. Detailed view on the five identified types of multiporous grooved sensilla using HR-SEM. Scale bar represents 500 nm. C. Detailed view on SSR traces of spontaneous spikes from the neocembrene-detecting sensillum of S I type, showing spikes of three different amplitudes. D. SSR traces recorded on neocembrene-detecting sensillum for neocembrene and geranylgeraniol.

The response spectrum of neocembrene sensillum to Panel 1 was markedly similar to that of ab3A neuron of PsimOR14-expressing Drosophila (Fig. 5A). None of the compounds elicited higher average responses that 10 Δ spikes/s, except for neocembrene and geranylgeraniol; their average Δ spikes/s were even slightly higher than those of heterologously expressed PsimOR14, reaching ∼65 and ∼29, respectively (Fig. 5A, B). Likewise, the dose-response experiment with neocembrene indicated a higher sensitivity threshold and lower ED50 = 0.016 ng (Fig. 5C, Supplementary Tables S10, S11).

SSR responses of the neocembrene-detecting SB I sensillum on the last flagellomere of P. simplex worker

A SSR responses to the main panel of volatiles with biological relevance for termites. The bars show the average Δ spikes/s values from 8–12 replicates ± SEM. The raw data is provided in Supplementary Tables S10. B. Characteristic SSR traces of the neocembrene-detecting SB I sensillum for neocembrene and geranylgeraniol. C. Dose response curve of the SSR responses to neocembrene by the SB I sensillum. The graph shows average Δ spikes/s values ± SEM based on 9–11 replicates. The curve fit and ED50 value were calculated using log(agonist) vs. response non-linear algorithm with least square fit method and the constraint of minimal response > 0. The raw data is provided in Supplementary Table S11.

PsimOR14 gene and protein structure, protein modeling, ligand docking, and MM-PBSA molecular dynamics (MD) simulations

Mapping the PsimOR14 transcript sequence on P. simplex draft genome revealed that the gene consists of six exons and is situated on the same locus and in close vicinity of PsimOR15, with which it shares the exon-intron boundaries, suggesting a recent diversification of the two genes via duplication, as also supported by their high sequence similarity (Fig. 6A, Fig. 1). Transcript (Fig. 6B) and protein (Fig. 6C) structures of PsimOR14 showed the presence of seven TMDs with the largest extracellular loop connecting TMD3 and TMD4 and the longest intracellular loop between TMD4 and TMD5.

PsimOR14 gene, transcript and protein structures, docking and MD simulations

A Genomic locus containing PsimOR14 and PsimOR15. PsimOR14 gene consists of 1 non-coding and 5 protein coding exons. B. PsimOR14 transcript with 6 exons represented as blue boxes with variable height according to protein-coding (higher) or untranslated (smaller) regions. ORF is marked by blue arrow. C. Transmembrane architecture of PsimOR14. In red are shown seven residues interacting with each of the three studied ligands, the green ellipse shows the intracellular flap the most impacted by ligand binding. D. Modelled apoform of PsimOR14. Red region denotes the binding site identified via docking, green region represents the intracellular flap region. E. Holoforms of PsimOR14 with sticks representing ligands without hydrogens (cyan = neocembrene, red = geranylgeraniol, orange = squalene). F. Relative mean volumes explored by atoms of PsimOR14 per simulation step when liganded with neocembrene, geranylgeraniol and squalene. PsimOR14 apoform represents the baseline zero. Gray lines show the binding site region, the green line delimits the intracellular flap the most affected by ligation. The whole PsimOR14 protein shown on the left, detail on the binding region in the middle and the intracellular flap on the right. Nucleotide and protein sequences of PsimOR14 are provided in Supplementary Table S12 and as NCBI entry under accession OR921181.

The initial PsimOR14 model was obtained using AlphaFold 2. After minimization and annealing of six molecular dynamics replicas, the structure with the lowest potential energy was selected for docking while assuming a homologous binding region to that reported in the structural characterization of MhOR5 from Machilis hrabei (7LIG) (del Mármol et al., 2021). The identified binding site, based on the superposition of 7LIG and our PsimOR14 model, is shown in Fig. 6D. For docking and MD simulations, three terpenoid ligands were selected from Panel 1, i.e., the best PsimOR14 agonists neocembrene and geranylgeraniol, and a weak agonist squalene (see Fig. 2). Their structures were parametrized using the CgenFF 4.6 forcefield via CHARMM-GUI and docked into the binding region of PsimOR14 using a flexible ligand docking algorithm. All the final binding poses occupy the same ligand binding site, and only the best-ranked pose for each ligand was selected for further analysis. Dockings are visualized in Fig. 6E, the final scores of the best-ranked poses are shown in Table 1. Based on the composition of the scores, we hypothesized that primarily Van der Waals interactions facilitate the binding, providing more possibilities to adapt a similar energy profile for the bound ligand.

Surprisingly, the ranking of the predicted binding strength squalene > geranylgeraniol > neocembrene did not correspond to the biological effect of the agonists observed in SSR assays (neocembrene > geranylgeranyl > squalene) (Table 1). Per-residue decomposition results (Supplementary Tables S13, S14) showed that all three ligands bind two hydrophobic patches made out of residues from TMD3 and TMD4 (Cys154, Ala157, Val158; Thr221, Leu224, Ala225, Tyr228). Neocembrene binds only these patches, while geranylgeraniol also interacts with Gly161, Phe162, Ile165, Ile217, and Val220. Squalene, being the largest ligand, additionally contacts Met37, Val166, and Phe169. Thus, the predicted binding strength order may be due to the increasing number of stabilizing contacts between the hydrophobic binding region and hydrophobic ligands, without being necessarily translated onto the biological effect on the receptor function based on the receptor selectivity.

By contrast, our subsequent analysis of PsimOR14 dynamic behaviour revealed an interesting impact of ligand binding on the dynamicity of the protein, especially in its intracellular region. Using a proprietary protocol and the convex hull algorithm, we analysed changes in per-atom dynamic behaviour across different forms. By tracing three-dimensional polytopes of atomic positions over time, we approximated the volume per step and arrived at a scalar Vex describing the total dynamicity of a region or the whole protein. As shown in Table 1 and Fig. 6F, the binding of the ligands has a strong effect on the dynamicity especially in the intracellular protein region manifested by relative dynamicity ordered as squalene > geranylgeranyl > neocembrene. Thus, the strength of the effect of different ligands on the intracellular region matches the functional SSR experiments and demonstrates the allosteric effect transferring the distribution of the binding to the protein periphery.

Caste-biased PsimOR14 expression and antennal sensitivity to neocembrene

We next decided to compare the expression pattern of PsimOR14 between P. simplex workers and soldiers, along with the sensitivity of the two castes to its preferred ligand, neocembrene. Differential expression analyses of the head RNAseq data from workers and soldiers revealed that PsimOR14 is significantly more expressed in workers, being among 10 of the most differentially expressed ORs (Fig. 7A, Supplementary Table S15). Subsequent EAG measurements were in line with this observation and indicated significantly stronger responses to neocembrene in workers (p=0.012) (Fig. 7B, Supplementary Table S16).

Caste comparison of PsimOR14 expression and EAG responses between P. simplex workers and soldiers

A Volcano plot representing edgeR differential gene expression analysis of all 50 P. simplex ORs in RNAseq data from soldier and workers heads. Colored dots mark ORs reaching absolute value of log2 fold change ≥ 1, horizontal lines represent p-value thresholds of 0.05 and 0.01. Numeric values of the edgeR and DESeq2 differential expression analysis are provided in Supplementary Table S15. B. EAG responses of whole antenna preparations of workers and soldiers to neocembrene at a dose of 10 ng (mean ± SD shown on log2 scale). Inter-caste differences were compared using t-test on log2-transformed data. Raw data is shown in Supplementary Table S16.

Discussion

Identification of PsimOR14 as a pheromone receptor narrowly tuned to neocembrene, the main component of the TFP in P. simplex, represents the first deorphanisation of an odorant receptor in termites. In contrast to eusocial Hymenoptera, the molecular aspects of termite chemical communication and orientation remain largely understudied despite the systematic effort previously dedicated to unveil the chemical identity, glandular origin and biological functions of termite pheromones (Bagnères & Hanus, 2015; Bordereau & Pasteels, 2011; Mitaka & Akino, 2021; Sillam-Dussès, 2010). Current knowledge is limited to a few studies listing the termite OR repertoire (Harrison et al., 2018; Johny et al., 2023; Terrapon et al., 2014), caste-specific OR expression (Mitaka et al., 2016), or addressing the significance of ORCo for olfactory function (Castillo et al., 2022; Gao et al., 2020).

Multiple evidence suggests that the complex communication and orientation demands in the colonies of eusocial Hymenoptera are facilitated by the greatly expanded repertoire of ORs, especially that of the 9-exon subfamily in ants and paper wasps participating in the detection of cuticular hydrocarbons (CHCs) as important cues in contact chemoreception of colony and caste identity and fertility status in eusocial insects (Engsontia et al., 2015; Legan et al., 2021; McKenzie et al., 2016; Pask et al., 2017; Zhou et al., 2015). The 9-exon subfamily and the overall OR richness has been inherited by the extant eusocial Hymenoptera from the ancestor of Aculeata (McKenzie et al., 2016), and also solitary aculeate taxa display large OR repertoires (Obiero et al., 2021); this preadaptation might had been important for the repeated emergence of eusociality and the related complex communication. This is supported by the convergent simplification of OR array (including 9-exon genes) in parasitic ant taxa, along with the simplification of their behavioral repertoire (Jongepier et al., 2022).

With their small numbers of ORs, termites clearly contradict the paradigm on eusociality as a driver of OR richness. In spite of the parsimony of their pheromone diversity, their chemical communication remains by far more complex than in solitary insects and includes pheromone components from a variety of chemical classes, including fatty-acyl derived alcohols and aldehydes, terpenoids, and, last, but not least, the CHCs. Independently of eusocial Hymenoptera, termites evolved an intricate communication system using CHCs as kin- and nestmate discrimination cues and as indicators of caste identity and fertility status (reviewed in Bagnères & Hanus, 2015; Mitaka & Akino, 2021). Therefore, the termite social evolution apparently adopted a different trajectory to accommodate the needs of chemical communication, including the detection of CHCs. One of alternative scenarios springs from observations that termites possess an extraordinarily rich set of ionotropic receptors (IRs), reaching up to more than one hundred IRs in some species (Harrison et al., 2018; Johny et al., 2023; Terrapon et al., 2014), and 95 IRs identified in P. simplex (Johny et al., 2023). Even though the richness of IRs is shared by termites and their cockroach relatives, with B. germanica having the largest IR repertoire ever identified in insects (Robertson et al., 2018), IRs underwent termite-specific expansions (Harrison et al., 2018; Johny et al., 2023). Since IRs have been shown to detect olfactory information (Benton et al., 2009), it has been speculated that the diversification of IRs may account for the complexity of termite chemical communication, as indirectly supported by caste and sex-specific pattern of their expression in termites (Harrison et al., 2018). This hypothesis remains to be experimentally verified.

Nevertheless, as we show here, the trail-following communication by means of TFP, having shared evolutionary origins with termite courtship communication (Bagnères & Hanus, 2015; Bordereau & Pasteels, 2011; Sillam-Dussès, 2010), is mediated by an odorant receptor. This is in line with the previous observations in two termite species showing that ORCo silencing impairs their ability to follow the foraging trail (Gao et al., 2020).

Neurophysiological characteristics of D. melanogaster ab3 neuron expressing PsimOR14 showed expected patterns of spontaneous firing rates of units of Hz (6.87 ± 4.73, mean ± SD) and maximum firing rates of 90 spikes/s at the highest used neocembrene doses (1 µg), though not reaching the reported maxima for ab3 responses with genuine and exogenous receptors, which may be well over 100 spikes/s (e.g., Chahda et al., 2019). This suggests that Drosophila Empty Neuron system co-expressing with the exogenous OR the Drosophila ORCo is a suitable technique for termite OR characterizations. The apparently narrow tuning of PsimOR14 to the cyclic diterpene hydrocarbon neocembrene, with receptor lifetime sparseness equal to 0.89, is typical for insect ORs adaptively shaped to detect pheromone components (pheromone receptors) or other volatiles with high biological importance, such as key food or host attractants (e.g., Carey et al., 2010; Fleischer & Krieger, 2018). With average Δ spike/s over 20, the linear diterpene alcohol geranylgeraniol was the only other agonist with non-negligible response; none of the remaining compounds elicited a Δ spike of more than 7. This ranking is interesting because the set of tested panels contained multiple other compounds derived from terpenoid scaffold, and invited to speculate about conformational similarities between neocembrene and geranylgeraniol. Even more interestingly, another related terpene alcohol, the monoterpene linalool, had the biggest negative Δ spike score (−6.8) and the absolute spikes/s ranged from 0 to 5 (1.5 ± 1.9, mean ± SD), suggesting a possible inverse agonist function of linalool.

The SSR-based identification of neocembrene-sensing sensillum in P. simplex workers provided a response pattern to Panel 1 very similar to that of PsimOR14-expressing ab3 Drosophila sensillum, e.g., strong response to neocembrene, followed by geranylgeraniol, supporting independently the results obtained from transgenic Drosophila. No other important responses to termite pheromone components from the Panel 1, which also includes the minor P. simplex TFP component (3Z,6Z,8E)-dodecatrien-1-ol, were recorded, in spite of the likely presence of two additional olfactory neurons harbored in the necembrene-detecting sensillum.

The basic PsimOR14 protein architecture showed a pattern typical for insect ORs, having seven TMDs. Likewise, the docking experiments with three selected ligands (neocembrene, geranylgeraniol and squalene) identified a binding site defined by a binding pocket deep in the transmembrane region, homologous to that in previously studied insect ORs, and confirmed the nature of the ligand binding to mainly rely on hydrophobic interactions (e.g., del Mármol et al., 2021; Pettersson & Cattaneo, 2023; Yuvaraj et al., 2021). In case of PsimOR14, mainly the hydrophobic and aromatic residues from TMD3 and TMD4 interacted with the studied ligands, as also suggested for some previously studied insect ORs (e.g., del Mármol et al., 2021; Yuvaraj et al., 2021).

Our computational analysis of PsimOR14 ligand binding did not provide a straightforward image of the binding affinity score as the best proxy of the biological effect size. The binding affinity, typically quantified through scoring functions that estimate the binding Gibbs free energy, often correlates well with experimental data, and may be useful for understanding how inhibitors might block native ligand binding. Yet, in systems where ligand binding induces allosteric effects, as is the case of insect ORs (del Mármol et al., 2021), the simple binding affinity does not provide a complete picture on receptor signal transduction function.

Our MM/PBSA analysis confirmed that the weakest agonist squalene exhibits the highest binding affinity, followed by geranylgeranyl and the strongest agonist neocembrene. This at the first view contradictory observation is in fact consistent with the number of stabilizing contacts identified for each ligand, which correlates with the space occupied by the ligands and increases from neocembrene through geranylgeraniol to squalene. By contrast, molecular dynamics analysis revealed that the binding of ligands reduces the dynamicity of PsimOR14’s intracellular region in the order neocembrene > geranylgeraniol > squalene. This dynamic effect aligns with the experimental data showing that neocembrene has the most significant biological impact. These results underscore the importance of considering both binding affinity and allosteric effects when analyzing ligand-receptor interactions.

PsimOR14 does not belong to the most expressed P. simplex ORs (Fig. 1). However, interestingly, it is among those having the most caste-biased expression with significantly higher transcript abundance in antennae of workers compared to soldiers. Accordingly, also the electrophysiological responses of workers to neocembrene were significantly larger in workers. Termite soldiers are known to lay pheromone trails, to detect TFPs and to participate in foraging and field exploration in a number of termite species, though the behavioral patterns of soldiers and workers during these activities differ (Kaib, 1990; Traniello, 1981; Traniello & Busher, 1985). This has also been demonstrated in a close relative to our model, the congeneric species Prorhinotermes inopinatus (Rupf & Roisin, 2008). Yet, differences in sensitivity of termite workers and soldiers have not previously been addressed at the electrophysiological level. Caste-biased PsimOR14 expression and neocembrene sensitivity may represent olfactory information filtering adaptive to the different tasks of the two castes, as documented, e.g., in ants (Caminer et al., 2023).

Genus Prorhinotermes is the most basally situated termite taxon having a terpenoid component as a part of its TFP (or SPP) (Sillam-Dussès et al., 2009). The acquisition of terpenoid pheromone components is undoubtedly due to the evolution of terpene biosynthesis in basal Neoisoptera or their ancestors, which is manifested by fascinating diversity of terpenoids produced by soldiers of Neoisoptera as defensive compounds (Gössinger, 2019). In line with general observations on pheromone evolution (Steiger et al., 2010), some of these terpenoid products of the soldier defensive frontal gland gained via exaptation a novel function of alarm pheromones (Dolejšová et al., 2014; Roisin et al., 1990; Šobotník et al., 2008, 2010), and the cyclic diterpenes became part of SPPs and TFPs in some species by co-opting the terpenoid biosynthesis in exocrine glands dedicated to SPP and TFP production. Beside the occurrence of neocembrene in the sub-basal Prorhinotermes, it only occurs as TFP and SPP component in much later diverging representatives of the modern family Termitidae (formerly also higher termites). It is particularly frequent as trail-following pheromone and sex-pairing pheromone in the diversified subfamily Nasutitermitinae (Bordereau & Pasteels, 2011). Because neocembrene is deemed to be the monocyclic precursor of the defensive polycyclic diterpenes produced by Nasutitermitinae soldiers, it is plausible that it gained the secondary pheromone role by expressing the responsible (unknown) terpene synthase in the pheromone glands (sternal gland of workers and soldiers for TFP, tergal and/or sternal glands of female imagoes for SPP). Once reliable genome or antennal transcriptome of Nasutitermitinae sequences become available, it would be of interest to search in these species for OR orthologs of PsimOR14 and test whether they have retained the function of neocembrene detection.

Within the available P. simplex OR sequences, the pheromone receptor responsible for the detection of (3Z,6Z,8E)-dodecatrien-1-ol should be the preferential target of future deorphanisation efforts. This compound became the most widespread component of termite TFPs and SPPs (Bordereau & Pasteels, 2011; Sillam-Dussès, 2010). It occurs in most lineages of Neoisoptera and is characteristic by extraordinary efficiency as evidenced by strikingly low activity threshold. In some species, 10-5 ng/cm of the trail is sufficient to elicit the trail-following behavior, while the activity thresholds for other known TFPs are by two or more orders of magnitude higher (Bordereau & Pasteels, 2011). In P. simplex, (3Z,6Z,8E)-dodecatrien-1-ol occurs not only as a minor component of TFP, but also as the main component of SPP, emitted by dispersing female in tiny quantities, which are sufficient to mediate the appropriate antennal and behavioral response of males.

An intuitive target for future functional analysis is PsimOR15, the likely tandem copy paralog of PsimOR14. Both genes are localized on the same locus, share the gene architecture, and their transcripts are equally represented in worker antennal transcriptome, though PsimOR15 does not display the caste-biased expression. PsimOR15 is thus a candidate for potential pheromone receptor function.

Materials and methods

Termites

Multiple laboratory colonies of P. simplex, originating from previous field collections in Cuba and Florida, are held in the Institute of Organic Chemistry and Biochemistry, Czech Academy of Sciences. Colonies are reared in glass vivaria at 27°C and 80% relative humidity in clusters of spruce wood slices. These colonies were used for RNA extraction, single sensillum recordings (SSR), electroantennographic recordings (EAG), and scanning electron microscopy (SEM). For all experiments, workers of 4th or 5th stage were selected as the most abundant developmental stages, recognized according to body size and head width.

RNA extraction, OR cloning and construct generation

Total RNA was extracted from 20 pairs of dissected P. simplex antennae using PureLink RNA Mini Kit (Invitrogen, Carlsbad, CA, USA) following the manufacturer’s protocol and quantified using NanoDrop spectrophotometer (Thermo, Delaware, USA). From the total RNA, 2 µg was used to synthesize the cDNA using SuperScript IV Reverse Transcriptase (Invitrogen, Carlsbad, CA, USA) according to manufacturer’s instructions. The efficiency of cDNA synthesis was evaluated by amplification of odorant receptor co-receptor (ORCo). The list of primers is provided in Supplementary Table S17).

The full-length open reading frame (ORF) of each selected PsimORs was PCR-amplified from the cDNA using the DreamTaq Green PCR Master Mix (Invitrogen, USA) and gene-specific primers (Supplementary Table S17). Amplification products were purified by QIAquick Gel Extraction Kit (Qiagen, Germany), cloned into pCR8/GW/TOPO vector using the TOPO TA Cloning Kit (Invitrogen, USA) and transformed into OneShot TOP10 competent cells (Invitrogen, USA). Positive colonies were selected based on colony-PCRs using primers GW1 and GW2, recombinant plasmids were isolated using the QIAprep 2.0 Spin Miniprep Columns (Qiagen, Germany) and sequences were verified by Sanger sequencing (Eurofins Genomics, Germany).

The expression vector constructs were prepared using the Gateway LR recombination cloning technology (Invitrogen, USA) based on recombination of the phage-like attachment sites attL/R in pCR8/GW/TOPO with the bacteria-like attachment site attB in pUASg.attb vector (obtained from Drosophila Genomics Resource Center, Bloomington, USA). The resulting constructs pUASg.attB-PsimOR were purified using the QIAprep 2.0 Spin Miniprep Columns (Qiagen, Germany) and insert sequences were verified by Sanger sequencing at Eurofins Genomics (Germany). All primers used for Sanger sequencing and colony-PCR are listed in Supplementary Table S17).

Fly lines

D. melanogaster lines used in the Empty Neuron system were kindly provided by Dr. Thomas O. Auer (from Richard Benton Lab, University of Lausanne, Switzerland). The wild type W1118line, used as a control, was kindly provided by Prof. Michal Žurovec (Biology Centre, Czechia). All Drosophila melanogaster lines were reared in an incubator which was set at 24±2°C with relative humidity of 50±5%. Flies were fed with in-house prepared diet based on standard cornmeal food. The fly lines used are listed in Supplementary Table S18.

Transgenic expression of termite ORs in D. melanogaster ab3A neuron

Selected PsimORs were expressed in the Drosophila melanogaster Empty Neuron system for functional screening. Transgenic D. melanogaster UAS-PsimOR lines were generated by BestGene Inc. (Chino Hills, CA, USA) by injecting pUASg.attB-PsimOR vectors into fly embryos expressing the integrase PhiC31 and carrying an attP landing site resulting in flies with genotype w−; +; UAS-PsimOR (w+)/+.

The recent CRISPR-Cas9-engineered empty neuron line Or22ab-Gal4(Chahda et al., 2019) was used as Δhalo genetic background for the expression of UAS-PsimOR in Dmel ab3 sensilla. The fly crossing scheme was adapted from (Gonzalez et al., 2016) with a modification at the F3 crossing. Final homozygote lines with UAS-PsimOR and Or22ab-Gal4 were generated and used for the electrophysiological recordings. The full description of the crossing scheme is provided in Supplementary Fig. S2.

Chemicals

For SSR measurements, we used a total of 73 chemicals organized into four panels. The main panel contained 17 compounds biologically relevant to termites, i.e. components of termite pheromones and cuticular hydrocarbons known from Neoisoptera and structurally related compounds. This panel was used for initial SSR screening of PsimOR9, 14, 30 and 31 in transgenic D. melanogaster and for SSR experiments with P. simplex workers. For detailed analysis of Psim OR14, three additional panels were used, consisting of 56 frequently occurring insect semiochemicals (e.g., terpenoids, fatty acid esters, fatty alcohols and aldehydes, etc.). Panel 1 compounds were diluted in n-hexane to 100 ng/µl, panel 2–4 compounds were diluted in paraffin oil to 10-3 v/v. List of all compounds tested and their origin are listed in Supplementary Table S19.

Organic synthesis

For the purpose of SSR experiments, we synthesized (Z)-dodec-3-en-1-ol, (3Z,6Z)-dodeca-3,6-dien-1-ol, (3Z,6Z,8E)-dodecatrien-1-ol, and dodec-3-yn-1-ol, and included these compounds into Panel 1. The synthesis of these fatty alcohols is described in the Supplementary Information file.

Electrophysiology

SSR recordings on Drosophila ab3 sensillum were performed as described previously (Benton & Dahanukar, 2023; Olsson & Hansson, 2013). We used 2–4 days old flies for one recording per each to avoid neuronal adaptations from multiple stimulations. To expose more ab3 sensilla, the fly preparation was done with arista down (Keesey et al., 2022). In termites, the olfactory sensilla situated on the last antennal flagellomere of workers was targeted for SSR, since their number increases towards the distal end of termite antennae, the last segment being significantly more populated by olfactory sensilla than any other segment (Castillo et al., 2021). The antennal preparation was done similarly as described for Drosophila with the grounding electrode inserted into the clypeus. The sensilla were observed under the Nikon FN1 eclipse microscope at 60× magnification. For all electrophysiological measurements, the recording electrode was brought into contact with the base of the sensillum using a Kleindiek Nanotechnik MM3A micromanipulator connected to a cubic micromanipulator device. Using Syntech stimulus delivery system (CS55 model, Syntech, Germany), the odorant stimulus was administered as a 0.3 s pulse by inserting the tip of the glass Pasteur pipette through a hole in a tube carrying a purified air stream (0.4 L/min). The distance between the antenna and odor delivery system was approximately 4 cm. From each diluted odorant (100 ng/µL), 10 µl were pipetted on 1 cm diameter filter paper disk placed in a glass Pasteur pipette in the screening experiment, while the doses ranging from 0.01 to 500 ng of neocembrene per filter paper were used in the dose-response experiments.

The signal was amplified and digitally converted using Syntech IDAC-4. The neuronal cells were sorted based on their amplitude and the spikes were counted using the AutoSpike v3.9software (Syntech Ockenfels, Germany). Δ spike was calculated by subtracting the number of spikes during 1 second post-stimulation from the number of spikes generated 1 second before the stimulation. In dose response experiments, the counting periods were 0.5 s. Δ spike values were corrected by subtracting the response generated by the solvent and converted into Δ spike/s (Benton & Dahanukar, 2023; Olsson & Hansson, 2013). The receptor lifetime sparseness value was calculated according to Chang et al. (Chang et al., 2023).

EAG experiment addressing the caste-specificity of antennal responses to neocembrene was performed with 15 workers and 15 soldiers; each individual was only used for one stimulation series consisting of air–hexane–neocembrene (10 ng)–hexane–air. Brain and antennal tip were placed between two Ag/AgCl electrodes containing Ringer’s solution and connected to a high impedance (1014 Ω) amplifier (Syntech, Buchenbach, Germany). The antennal preparation was placed into a stream of cleaned air (500 mL/min), into which the stimuli were injected from Pasteur pipettes containing a 1.5 cm2 filter paper impregnated with 10 µL of the tested solution. Odor injections were controlled by a foot switch-operated Syntech stimulus controller and maximal negative deflection was measured using Syntech EagPro software. Pasteur pipettes containing odorant stimuli were changed after three stimulations. Air responses were used for data normalization, the responses log2-transformed to reduce heteroscedasticity and comply with assumptions for parametric testing (Bartlett test for equal variances, and Shapiro-Wilk normality test), and then compared between workers and soldiers using Student’s t-test.

Scanning electron microscopy

For SEM, 10 workers with intact antennae were cold-anesthetized and decapitated with micro-scissors. Heads were desiccated in increasing ethanol concentrations (60, 80, 90, and 96%, each for 2 h) followed by 12 h in acetone. Heads were then attached to aluminium holders for microscopy using adhesive tape, and differentially oriented to allow axial, dorsal, ventral, and lateral views. The samples were gold-coated for regular SEM (4 nm gold layer) and high-resolution SEM (HR-SEM, 2 nm) using sputter coater Bal-Tec SCD 050. Last antennal segments were inspected and photographed under scanning electron microscope JEOL 6380 LV (SEM). Surface of particular sensilla were studied using high resolution field emission scanning electron microscope JSM-IT800 (HR-SEM) and Olympus Soft Imaging Solution software. Working distance for all samples was 4.0–4.1 mm and accelerating voltage 2.0 kV.

Bioinformatics

For phylogenetic reconstruction of termite ORs, we used 182 OR protein sequences originating from five termite species, i.e. Neotermes cubanus, Prorhinotermes simplex, Inquilinitermes inquilinus (Johny et al., 2023), Zootermopsis nevadensis (Terrapon et al., 2014) and Cryptotermes secundus (Harrison et al., 2018), and the bristletail Lepisma saccharina (Thoma et al., 2019) as a basal insect outgroup. For all species also the ORCo sequence was included. Sequences were aligned by means of the MUSCLE algorithm and used for reconstruct the phylogenetic tree with the IQ-TREE maximum likelihood algorithm (Nguyen et al., 2015) using the JTT+F+R8 substitution model and 10,000 ultrafast bootstrap replicates.

The gene structures of PsimORs were characterized by local alignment of full-length transcript sequences from Johny et al. (Johny et al., 2023) to our in-house genome assembly using BLAST (for details on genome assembly see Koubová et al., 2021) and confirmed with genomic mapping of the RNAseq data from P. simplex antennae available under accession SRX17749141 in NCBI SRA archives using STAR aligner v2.7.10b (Dobin et al., 2013). The mapping results were further used for abundance estimations of all ORs and ORCo in antennal transcriptome reported in Johny et al. (Johny et al., 2023). Read counts were obtained with featureCounts tool from the Subread package (https://subread.sourceforge.net/) and normalized according to the FPKM (Fragments Per Kilobase Million) method.

Differential OR expression analysis in P. simplex soldier and worker heads was performed using the RNAseq data available in SRA archives under accessions SRX18952230-32 and SRX18952237-39. Read counts obtained using STAR mapping and featureCounts estimations were statistically evaluated using the DESeq2 Bioconductor package in R and edgeR.

PsimOR14 secondary structure was predicted using online tools Jpred 4.0.0 (www.compbio.dundee.ac.uk/jpred) and TMHMM2.0 (https://services.healthtech.dtu.dk/services/TMHMM-2.0), schematic model was generated using Protter (https://wlab.ethz.ch/protter).

Supporting information

eLife Assessment

This important study presents the first identification of the odorant receptor for the trail pheromone in termites. The evidence supporting the conclusions is compelling, with state-of-the-art neurophysiological and genetic methods. The work will be of broad interest in multiple disciplines, such as entomology, chemical ecology, and sensory physiology.

Reviewer #1 (Public review):

Summary:

In their comprehensive analysis Diallo et al. deorphanise the first olfactory receptor of a non-hymenopteran eusocial insect - a termite and identified the well-established trail pheromone neocembrene as the receptor's best ligand. By using a large set of odorants the authors convincingly show that, as expected for a pheromone receptor, PsimOR14 is very narrowly tuned. While the authors first make use of an ectopic expression system, the empty neuron of Drosophila melanogaster, to characterise the receptor's responses, they next perform single sensillum recordings with different sensilla types on the termite antenna. By that, they are able to identify a sensillum that houses three neurons, of which the B neuron exhibits the narrow responses described for PsimOR14. Hence the authors do not only identify the first pheromone receptor in a termite but can even localize its expression on the antenna. The authors in addition perform a structural analysis to explain the binding properties of the receptor and its major and minor ligands (as this is beyond my expertise, I cannot judge this part of the manuscript). Finally, they compare expression patterns of ORs in different castes and find that PsimOR14 is more strongly expressed in workers than in soldier termites, which corresponds well with stronger antennal responses in the worker caste.

Strengths:

The manuscript is well-written and a pleasure to read. The figures are beautiful and clear. I actually had a hard time coming up with suggestions.

Weaknesses:

Whenever it comes to the deorphanization of a receptor and its potential role in behaviour (in the case of the manuscript it would be trail-following of the termite) one thinks immediately of knocking out the receptor to check whether it is necessary for the behaviour. However, I definitely do not want to ask for this (especially as the establishment of CRISPR Cas-9 in eusocial insects usually turns out to be a nightmare). I also do not know either, whether knockdowns via RNAi have been established in termites, but maybe the authors could consider some speculation on this in the discussion.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors performed the functional analysis of odorant receptors (ORs) of the termite Prorhinotermes simplex to identify the receptor of trail-following pheromone. The authors performed single-sensillum recording (SSR) using the transgenic Drosophila flies expressing a candidate of the pheromone receptor and revealed that PsimOR14 strongly responds to neocembrene, the major component of the pheromone. Also, the authors found that one sensillum type (S I) detects neocembrene and also performed SSR for S I in wild termite workers. Furthermore, the authors revealed the gene, transcript, and protein structures of PsimOR14, predicted the 3D model and ligand docking of PsimOR14, and demonstrated that PsimOR14 is higher expressed in workers than soldiers using RNA-seq for heads of workers and soldiers of P. simplex and that EAG response to neocembrene is higher in workers than soldiers. I consider that this study will contribute to further understanding of the molecular and evolutionary mechanisms of the chemoreception system in termites.

Strength:

The manuscript is well written. As far as I know, this study is the first study that identified a pheromone receptor in termites. The authors not only present a methodology for analyzing the function of termite pheromone receptors but also provide important insights in terms of the evolution of ligand selectivity of termite pheromone receptors.

Weakness:

As you can see in the "Recommendations to the Authors" section below, there are several things in this paper that are not fully explained about experimental methods. Except for this point, this paper appears to me to have no major weaknesses.

Reviewer #3 (Public review):

Summary:

Chemical communication is essential for the organization of eusocial insect societies. It is used in various important contexts, such as foraging and recruiting colony members to food sources. While such pheromones have been chemically identified and their function demonstrated in bioassays, little is known about their perception. Excellent candidates are the odorant receptors that have been shown to be involved in pheromone perception in other insects including ants and bees but not termites. The authors investigated the function of the odorant receptor PsimOR14, which was one of four target odorant receptors based on gene sequences and phylogenetic analyses. They used the Drosophila empty neuron system to demonstrate that the receptor was narrowly tuned to the trail pheromone neocembrene. Similar responses to the odor panel and neocembrene in antennal recordings suggested that one specific antennal sensillum expresses PsimOR14. Additional protein modeling approaches characterized the properties of the ligand binding pocket in the receptor. Finally, PsimOR14 transcripts were found to be significantly higher in worker antennae compared to soldier antennae, which corresponds to the worker's higher sensitivity to neocembrene.

Strengths:

The study presents an excellent characterization of a trail pheromone receptor in a termite species. The integration of receptor phylogeny, receptor functional characterization, antennal sensilla responses, receptor structure modeling, and transcriptomic analysis is especially powerful. All parts build on each other and are well supported with a good sample size.

Weaknesses:

The manuscript would benefit from a more detailed explanation of the research advances this work provides. Stating that this is the first deorphanization of an odorant receptor in a clade is insufficient. The introduction primarily reviews termite chemical communication and deorphanization of olfactory receptors previously performed. Although this is essential background, it lacks a good integration into explaining what problem the current study solves.

Selecting target ORs for deorphanization is an essential step in the approach. Unfortunately, the process of choosing these ORs has not been described. Were the authors just lucky that they found the correct OR out of the 50, or was there a specific selection process that increased the probability of success?

The authors assigned antennal sensilla into five categories. Unfortunately, they did not support their categories well. It is not clear how they were able to differentiate SI and SII in their antennal recordings.

The authors used a large odorant panel to determine receptor tuning. The panel included volatile polar compounds and non-volatile non-polar hydrocarbons. Usually, some heat is applied to such non-volatile odorants to increase volatility for receptor testing. It is unclear how it is possible that these non-volatile compounds can reach the tested sensilla without heat application.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In their comprehensive analysis Diallo et al. deorphanise the first olfactory receptor of a nonhymenopteran eusocial insect - a termite and identified the well-established trail pheromone neocembrene as the receptor's best ligand. By using a large set of odorants the authors convincingly show that, as expected for a pheromone receptor, PsimOR14 is very narrowly tuned. While the authors first make use of an ectopic expression system, the empty neuron of Drosophila melanogaster, to characterise the receptor's responses, they next perform single sensillum recordings with different sensilla types on the termite antenna. By that, they are able to identify a sensillum that houses three neurons, of which the B neuron exhibits the narrow responses described for PsimOR14. Hence the authors do not only identify the first pheromone receptor in a termite but can even localize its expression on the antenna. The authors in addition perform a structural analysis to explain the binding properties of the receptor and its major and minor ligands (as this is beyond my expertise, I cannot judge this part of the manuscript). Finally, they compare expression patterns of ORs in different castes and find that PsimOR14 is more strongly expressed in workers than in soldier termites, which corresponds well with stronger antennal responses in the worker caste.

Strengths:

The manuscript is well-written and a pleasure to read. The figures are beautiful and clear. I actually had a hard time coming up with suggestions.

We thank the reviewer for the positive comments.

Weaknesses:

Whenever it comes to the deorphanization of a receptor and its potential role in behaviour (in the case of the manuscript it would be trail-following of the termite) one thinks immediately of knocking out the receptor to check whether it is necessary for the behaviour. However, I definitely do not want to ask for this (especially as the establishment of CRISPR Cas-9 in eusocial insects usually turns out to be a nightmare). I also do not know either, whether knockdowns via RNAi have been established in termites, but maybe the authors could consider some speculation on this in the discussion.

We agree that a functional proof of the PsimOR14 function using reverse genetics would be a valuable addition to the study to firmly establish its role in trail pheromone sensing. Nevertheless, such a functional proof is difficult to obtain. Due to the very slow ontogenetic development inherent to termites (several months from an egg to the worker stage) the CRISPR Cas-9 is not a useful technique for this taxon. By contrast, termites are quite responsive to RNAimediated silencing and RNAi has previously been used for the silencing of the ORCo co-receptor in termites resulting in impairment of the trail-following behavior (DOI: 10.1093/jee/toaa248). Likewise, our previous experiments showed a decreased ORCo transcript abundance, lower sensitivity to neocembrene and reduced neocembrene trail following upon dsPsimORCo administration to P. simplex workers, while we did not succeed in reducing the transcript abundance of PsimOR14 upon dsPsimOR14 injection. We do not report these negative results in the present manuscript so as not to dilute the main message. In parallel, we are currently developing an alternative way of dsRNA delivery using nanoparticle coating, which may improve the RNAi experiments with ORs in termites.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors performed the functional analysis of odorant receptors (ORs) of the termite Prorhinotermes simplex to identify the receptor of trail-following pheromone. The authors performed single-sensillum recording (SSR) using the transgenic Drosophila flies expressing a candidate of the pheromone receptor and revealed that PsimOR14 strongly responds to neocembrene, the major component of the pheromone. Also, the authors found that one sensillum type (S I) detects neocembrene and also performed SSR for S I in wild termite workers. Furthermore, the authors revealed the gene, transcript, and protein structures of PsimOR14, predicted the 3D model and ligand docking of PsimOR14, and demonstrated that PsimOR14 is higher expressed in workers than soldiers using RNA-seq for heads of workers and soldiers of P. simplex and that EAG response to neocembrene is higher in workers than soldiers. I consider that this study will contribute to further understanding of the molecular and evolutionary mechanisms of the chemoreception system in termites.

Strength:

The manuscript is well written. As far as I know, this study is the first study that identified a pheromone receptor in termites. The authors not only present a methodology for analyzing the function of termite pheromone receptors but also provide important insights in terms of the evolution of ligand selectivity of termite pheromone receptors.

We thank the reviewer for the overall positive evaluation of the manuscript.

Weakness:

As you can see in the "Recommendations to the Authors" section below, there are several things in this paper that are not fully explained about experimental methods. Except for this point, this paper appears to me to have no major weaknesses.

We address point by point the specific comments listed in the Recommendation to the authors chapter below.

Reviewer #3 (Public review):

Summary:

Chemical communication is essential for the organization of eusocial insect societies. It is used in various important contexts, such as foraging and recruiting colony members to food sources. While such pheromones have been chemically identified and their function demonstrated in bioassays, little is known about their perception. Excellent candidates are the odorant receptors that have been shown to be involved in pheromone perception in other insects including ants and bees but not termites. The authors investigated the function of the odorant receptor PsimOR14, which was one of four target odorant receptors based on gene sequences and phylogenetic analyses. They used the Drosophila empty neuron system to demonstrate that the receptor was narrowly tuned to the trail pheromone neocembrene. Similar responses to the odor panel and neocembrene in antennal recordings suggested that one specific antennal sensillum expresses PsimOR14. Additional protein modeling approaches characterized the properties of the ligand binding pocket in the receptor. Finally, PsimOR14 transcripts were found to be significantly higher in worker antennae compared to soldier antennae, which corresponds to the worker's higher sensitivity to neocembrene.

Strengths:

The study presents an excellent characterization of a trail pheromone receptor in a termite species. The integration of receptor phylogeny, receptor functional characterization, antennal sensilla responses, receptor structure modeling, and transcriptomic analysis is especially powerful. All parts build on each other and are well supported with a good sample size.

We thank the reviewer for these positive comments.

Weaknesses:

The manuscript would benefit from a more detailed explanation of the research advances this work provides. Stating that this is the first deorphanization of an odorant receptor in a clade is insufficient. The introduction primarily reviews termite chemical communication and deorphanization of olfactory receptors previously performed. Although this is essential background, it lacks a good integration into explaining what problem the current study solves.

We understand the comment about the lack of an intelligible cue to highlight the motivation and importance of the present study. In the current version of the manuscript the introduction has been reworked. As suggested by Reviewer 3 in the Recommendations section below, the introduction now integrates some parts of the original discussion, especially the part discussing the OR evolution and emergence of eusociality in hymenopteran social insects and in termites, while underscoring the need of data from termites to compare the commonalities and idiosyncrasies in neurophysiological (pre)adaptations potentially linked with the independent eusociality evolution in the two main social insect clades.

Selecting target ORs for deorphanization is an essential step in the approach. Unfortunately, the process of choosing these ORs has not been described. Were the authors just lucky that they found the correct OR out of the 50, or was there a specific selection process that increased the probability of success?

Indeed, we were extremely lucky. Our strategy was to first select a modest set of ORs to confirm the feasibility of the Empty Neuron Drosophila system and newly established SSR setup, while taking advantage of having a set of termite pheromones, including those previously identified in the P. simplex model, some of them de novo synthesized for this project. The selection criteria for the first set of four receptors were (i) to have full-length ORF and at least 6 unambiguously predicted transmembrane regions, and (ii) to be represented on different branches (subbranches) of the phylogenetic tree. Then it was a matter of a good luck to hit the PsimOR14 selectively responding to the genuine P. simplex trail-following pheromone main component. In the revised version, we state these selection criteria in the results section (Phylogenetic reconstruction and candidate OR selection).

The deorphanization attempts of additional P. simplex ORs are currently running.

The authors assigned antennal sensilla into five categories. Unfortunately, they did not support their categories well. It is not clear how they were able to differentiate SI and SII in their antennal recordings.

We agree that the classification of multiporous sensilla into five categories lacks robust discrimination cues. The identification of the neocembrene-responding sensillum was initially carried out by SSR measurements on individual olfactory sensilla of P. simplex workers one-by-one and the topology of each tested sensillum was recorded on optical microscope photographs taken during the SSR experiment. Subsequently, the SEM and HR-SEM were performed in which we localized the neocembrene sensillum and tried to find distinguishing characters. We admit that these are not robust. Therefore, in the revised version of the manuscript we decided to abandon the attempt of sensilla classification and only report the observations about the specific sensillum in which we consistently recorded the response to neocembrene (and geranylgeraniol). The modifications affect Fig. 4, its legend and the corresponding part of the results section (Identification of P. simplex olfactory sensillum responding to neocembrene).

The authors used a large odorant panel to determine receptor tuning. The panel included volatile polar compounds and non-volatile non-polar hydrocarbons. Usually, some heat is applied to such non-volatile odorants to increase volatility for receptor testing. It is unclear how it is possible that these non-volatile compounds can reach the tested sensilla without heat application.

The reviewer points at an important methodological error we made while designing the experiments. Indeed, the inclusion of long-chain hydrocarbons into Panel 1 without additional heat applied to the odor cartridges was inappropriate, even though the experiments were performed at 25–26 °C. We carefully considered the best solution to correct the mistake and finally decided to remove all tested ligands beyond C22 from Panel 1, i.e. altogether five compounds. These changes did not affect the remaining Panels 2-4 (containing compounds with sufficient volatility), nor did they affect the message of the manuscript on highly selective response of PsimOR14 to neocembrene (and geranylgeryniol). In consequence, Figures 2, 3 and 5 were updated, along with the supplementary tables containing the raw data on SSR measurements. In addition, the tuning curve for PsimOR14 was re-built and receptor lifetime sparseness value re-calculated (without any important change). We also exchanged squalene for limonene in the docking and molecular dynamics analysis and made new calculations.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) L 208: "than" instead of "that"

Corrected.

(2) L 527+527 strange squares (•) before dimensions

Apparently an error upon file conversion, corrected.

(3) L553 "reconstructing" instead of "reconstruct"

Corrected.

(4) Two references (Chahda et al. and Chang et al. appear too late in the alphabet.

Corrected. Thank you for spotting this mistake. Due to our mistake the author list was ordered according to the alphabet in Czech language, which ranks CH after H.

Reviewer #2 (Recommendations for the authors):

(1) L148: Why did the authors select only four ORs (PsimOR9, 14, 30, and 31) though there are 50 ORs in P. simplex? I would like you to explain why you chose them.

Our strategy was to first select a modest set of ORs to confirm the feasibility of the Empty Neuron Drosophila system and newly established SSR setup, while taking advantage of having a set of termite pheromones, including those previously identified in the P. simplex model, some of them de novo synthesized for this project. Then, it was a matter of a good luck to hit the PsimOR14 selectively responding to the genuine P. simplex trail-following pheromone main component, while the deorphanization attempts of a set of additional P. simplex ORs is currently running. In the revised version of the manuscript, we state the selection criteria for the four ORs studied in the Results section (Phylogenetic reconstruction and candidate OR selection).

(2) L149: Where is Figure 1A? Does this mean Figure 1?

Thank you for spotting this mistake. Fig. 1 is now properly labelled as Fig. 1A and 1B in the figure itself and in the legend. Also the text now either refers to either 1A or 1B.

(3) Figure 1: The authors also showed the transcription abundance of all 50 ORs of P. simplex in the right bottom of Figure 1, but there is no explanation about it in the main text.

The heatmap reporting the transcript abundances is now labelled as Fig. 1B and is referred to in the discussion section (in the original manuscript it was referred to on the same place as Fig. 1).

(4) L260-265: The authors confirmed higher expression of PsimOR14 in workers than soldiers by using RNA-seq data and stronger EAG responses of PsimOR14 to neocembrene in workers than soldiers, but I think that confirming the expression levels of PsimOR14 in workers and soldiers by RT-qPCR would strengthen the authors' argument (it is optional).

qPCR validation is a suitable complement to read count comparison of RNA Seq data, especially when the data comes from one-sample transcriptomes and/or low coverage sequencing. Yet, our RNA Seq analysis is based on sequencing of three independent biological replicates per phenotype (worker heads vs. soldier heads) with ~20 millions of reads per sample. Thus, the resulting differential gene expression analysis is a sufficient and powerful technique in terms of detection limit and dynamic range.

We admit that the replicate numbers and origin of the RNA seq data should be better specified since the Methods section only referred to the GenBank accession numbers in the original manuscript. Therefore, we added more information in the Methods section (Bioinformatics) and make clear in the Methods that this data comes from our previous research and related bioproject.

(5) L491: I think that "The synthetic processes of these fatty alcohols are ..." is better.

We replaced the sentence with “The de novo organic synthesis of these fatty alcohols is described …”

(6) L525 and 527: There are white squares between the number and the unit. Perhaps some characters have been garbled.

Apparently an error upon file conversion, corrected.

(7) L795: ORCo?

Corrected.

(8) L829-830 & Figure 4: Where is Figure 4D?

Thank you for spotting this mistake from the older version of Figure 4. The SSR traces referred to in the legend are in fact a part of Figure 5. Moreover, Figure 4 is now reworked based on the comments by Reviewer 3.

(9) L860-864: Why did the authors select the result of edgeR for the volcano plot in Figure 7 although the authors use both DESeq2 and edgeR? An explanation would be needed.

Both algorithms, DESeq2 and EdgeR, are routinely used for differential gene expression analysis. Since they differ in read count normalization method and statistical testing we decided to use both of them independently in order to reduce false positives. Because the resulting fold changes were practically identical in both algorithms (results for both analyses are listed in Supplementary table S15), we only reported in Fig. 7 the outputs for edgeR to avoid redundancies. We added in the Results section the information that both techniques listed PsimOR14 among the most upregulated in workers.

Reviewer #3 (Recommendations for the authors):

The discussion contains many descriptions that would fit better into the introduction, where they could be used to hint at the study's importance (e.g., 292-311, 381-412). The remaining parts often lack a detailed discussion of the results that integrates details from other insect studies. Although references were provided, no details were usually outlined. It would be helpful to see a stronger emphasis on what we learn from this study.

Along with rewriting the introduction, we also modified the discussion. As suggested, the lines 292-311 were rewritten and placed in the introduction. By contrast, we preferred to keep the two paragraphs 381-412 in the discussion, since both of them outline the potential future interesting targets of research on termite ORs.

As suggested, the discussion has been enriched and now includes comparative examples and relevant references about the broad/narrow selectivity of insect ORs, about the expected breadth of tuning of pheromone receptors vs. ORs detecting environmental cues, about the potential role of additional neurons housed in the neocembrene-detecting sensillum of P. simplex workers, etc. From both introduction and discussion the redundant details on the chemistry of termite communication have been removed.

This includes explanations of the advantages of the specific methodologies the authors used and how they helped solve the manuscript's problem. What does the phylogeny solve? Was it used to select the ORs tested? It would be helpful to discuss what the phylogeny shows in comparison to other well-studied OR phylogenies, like those from the social Hymenoptera.

We understand the comment. In fact, our motivation to include the phylogenetic tree of termite ORs was essentially to demonstrate (i) the orthologous nature of OR diversity with few expansions on low taxonomic levels, and (ii) to demonstrate graphically the relationship among the four selected sequences. We do not attempt here for a comprehensive phylogenetic analysis, because it would be redundant given that we recently published a large OR phylogeny which includes all sequences used in the present manuscript and analysed them in the proper context of related (cockroaches) and unrelated insect taxa (Johny et al., 2023). This paper also discusses the termite phylogenetic pattern with those observed in other Insecta. This paper is repeatedly cited on appropriate places of the present manuscript and its main observations are provided in the Introduction section. Therefore, we feel that thorough discussion on termite phylogeny would be redundant in the present paper.

The authors categorized the sensilla types. Potential problems in the categorization aside, it would be helpful to know if it is expected that you have sensilla specialized in perceiving one specific pheromone. What is known about sensilla in other insects?

We understand. In the discussion of the revised version, we develop more about the features typical/expected for a pheromone receptor and the sensillum housing this receptor together with two other olfactory sensory neurons, including examples from other insects.

As the manuscript currently stands, specialist readers with their respective background knowledge would find this study very interesting. In contrast, the general reader would probably fail to appreciate the importance of the results.

We hope that the re-organized and simplified introduction may now be more intelligible even for non-specialist readers.

(1) L35: Should "workers" be replaced with "worker antennae"?

Corrected.

(2) L62: Should "conservativeness" be replaced by "conservation"?

Replaced with “parsimony”.

(3) L129: How and why did the authors choose four candidate ORs? I could not find any information about this in the manuscript. I wondered why they did not pick the more highly expressed PsimOr20 and 26 (Figure 7).

As already replied above in the Weaknesses section, we selected for the first deorphanization attempts only a modest set of four ORs, while an additional set is currently being tested. We also explained above the inclusion criteria, i.e. (i) full-length ORF and at least 6 unambiguously predicted transmembrane regions, and (ii) presence on different branches (subbranches) of the OR phylogeny. For these reasons, we did not primarily consider the expression patterns of different ORs. As for Fig. 7, it shows differential expression between soldiers and workers, which was not the primary guideline either and the data was obtained only after having the ORs tested by SSR. Yet, even though we had data on P. simplex ORs expression (Fig. 1B), we did not presume that pheromone receptors should be among the most expressed ORs, given the richness of chemical cues detected by worker termites and unlike, e.g., male moths, where ORs for sex pheromones are intuitively highly expressed.

The strategy of OR selection is specified in the results section of the revised manuscript under “Phylogenetic reconstruction and candidate OR selection”.

(4) 198 to 200: SI, II, and III look very similar. Additional measurements rather than qualitative descriptions are required to consider them distinct sensilla. The bending of SIII could be an artifact of preparation. I do not see how the authors could distinguish between SI and SII under the optical microscope for recordings. A detailed explanation is required.

As we responded above in “Weaknesses” chapter, we admit that the sensilla classification is not intelligible. Therefore, we decided in the revised version to abandon the classification of sensilla types and only focus on the observations made on the neocembreneresponding sensillum. To recognize the specific sensillum, we used its topology on the last antennal segment. Because termite antennae are not densely populated with sensilla, it is relatively easy to distinguish individual sensilla based on their topology on the antenna, both in optical microscope and SEM photographs. The modifications affect Fig. 4, its legend and the corresponding part of the results section (Identification of P. simplex olfactory sensillum responding to neocembrene).

(5) 208: "Than" instead of "that"

Corrected.

(6) 280: I suggest replacing "demand" with "capabilities"

Corrected.

(7) 312: Why "nevertheless? It sounds as if the authors suggest that there is evidence that ORs are not important for communication. This should be reworded.

We removed “Nevertheless” from the beginning of the sentence.

(8) 321 to 323: This sentence sounds as if something is missing. I suggest rewriting it.

This sentence simply says that empty neuron Drosophila is a good tool for termite OR deorphanization and that termite ORs work well Drosophila ORCo. We reworded the sentence.

(9) 323: I suggest starting a new paragraph.

Corrected.

(10) 421: How many colonies were used for each of the analyses?

The data for this manuscript were collected from three different colonies collected in Cuba. We now describe in the Materials and Methods section which analyses were conducted with each of the colonies.

(11) 430: Did the termites originate from one or multiple colonies and did the authors sample from the Florida and Cuba population?

The data for this manuscript were collected from three different colonies collected in Cuba. We now describe in the Materials and Methods section which analyses were conducted with each of the colonies.

(12) 501: How was the termite antenna fixated? The authors refer to the Drosophila methods, but given the large antennal differences between these species, more specific information would be helpful.

Understood. We added the following information into the Methods section under “Electrophysiology”: “The grounding electrode was carefully inserted into the clypeus and the antenna was fixed on a microscope slide using a glass electrode. To avoid the antennal movement, the microscope slide was covered with double-sided tape and the three distal antennal segments were attached to the slide.”

(13)509: I want to confirm that the authors indicate that the outlet of the glass tube with the airstream and odorant is 4 cm away from the Drosophila or termite antenna. The distance seems to be very large.

Thank you for spotting this obvious mistake. The 4 cm distance applies for the distance between the opening for Pasteur pipette insertion into the delivery tube, the outlet itself is situated approx. 1 cm from the antenna. This information is now corrected.

(14) 510/527: It looks like all odor panels were equally applied onto the filter paper despite the difference in solvent (hexane and paraffin oil). How was the solvent difference addressed?

In our study we combine two types of odorant panels. First, we test on all four studied receptors a panel containing several compounds relevant for termite chemical communication including the C12 unsaturated alcohols, the diterpene neocembrene, the sesquiterpene (3R,6E)-nerolidol and other compounds. These compounds are stored in the laboratory as hexane solutions to prevent the oxidation/polymerization and it is not advisable to transfer them to another solvent. In the second step we used three additional panels of frequently occurring insect semiochemicals, which are stored as paraffin oil solutions, so as to address the breadth of PsimOR14 tuning. We are aware that the evaporation dynamics differ between the two solvents but we did not have any suitable option how to solve this problem. We believe that the use of the two solvents does not compromise the general message on the receptor specificity. For each panel, the corresponding solvent is used as a control. Similarly, the use of two different solvents for SSR can be encountered in other studies, e.g. 10.1016/j.celrep.2015.07.031.

(15) 518: delta spikes/sec works for all tables except for the wild type in Table S5. I could not figure out how the authors get to delta spikes/sec in that table.

Thank you for your sharp eye. Due to our mistake, the values of Δ spikes per second reported in Table S5 for W1118 were erroneously calculated using the formula for 0.5 sec stimulation instead of 1 sec. We corrected this mistake which does not impact the results interpretation in Table S5 and Fig. 2.

522: Did the workers and soldiers originate from different colonies or different populations?

We now clearly describe in the Material and Methods section the origin of termites for different experiments. EAG measurements were made using individuals (workers, soldiers) from one Cuban colony.

(16) Figure 6C/D: I suggest matching colors between the two figures. For example, instead of using an orange circle in C and a green coloration of the intracellular flap in D, I recommend using blue, which is not used for something else. In addition, the binding pocket could be separated better from anything else in a different color.

We agree that the color match for the intracellular flap was missing. This figure is now reworked and the colors should have a better match and the binding region is better delineated.

(17) Figure 7/Table S15: It is unclear where the transcriptome data originate and what they are based on. Are these antennal transcriptomes or head transcriptomes? Do these data come from previous data sets or data generated in this study? Figure 7 refers to heads, Table S15 to workers and soldiers, and the methods only refer to antennal extractions. This should be clarified in the text, the figure, and the table.

We admit that the replicate numbers and origin of the RNA seq data should be better specified and that the information that the RNASeq originated from samples of heads+antennae of workers and soldiers should be provided at appropriate places. Therefore, we added more information on replicates and origin of the data in the Methods section (Bioinformatics) and make clear that this data comes from our previous research and refer to the corresponding bioproject. Likewise, the Figure 7 legend and Table S15 heading have been updated.

Version 2

Abstract

Abstract

Chemical communication is the cornerstone of eusocial insect societies since it mediates the social hierarchy, division of labor, and concerted activities of colony members. The chemistry of social insect pheromones received considerable attention in both major groups of social insects, the eusocial Hymenoptera and termites. By contrast, current knowledge on molecular mechanisms of social insect pheromone detection by odorant receptors (ORs) is limited to hymenopteran social insects and no OR was yet functionally characterized in termites, the oldest eusocial insect clade. Here, we present the first OR deorphanization in termites. We selected four OR sequences from the previously annotated antennal transcriptome of the termite Prorhinotermes simplex (Rhinotermitidae), expressed them in Empty Neuron Drosophila, and functionally characterized using single sensillum recording (SSR) and a panel of termite semiochemicals. In one of the selected ORs, PsimOR14, we succeeded in obtaining strong and reliable responses to the main component of P. simplex trail-following pheromone, the monocyclic diterpene neocembrene. PsimOR14 showed a narrow tuning to neocembrene; only one additional compound out of 67 tested (geranylgeraniol) generated non-negligible responses. Subsequently, we used SSR in P. simplex workers and identified the olfactory sensillum specifically responding to neocembrene, thus likely expressing PsimOR14. We report on homology-based modelling of neocembrene binding by PsimOR14 and show how different ligands impact the receptor dynamicity using molecular dynamics simulations. Finally, we demonstrate that PsimOR14 is significantly more expressed in worker antennae compared to soldiers, which correlates with higher sensitivity of workers to neocembrene.

Introduction

Chemical communication is the cornerstone of eusocial insect societies. It mediates the social hierarchy and division of labor, social cohesion and concerted activities of colony members. Different clades of eusocial insects have independently evolved chemical signals acting in similar social contexts, such as trail, alarm and queen pheromones, colony and caste recognition cues, and other signals. The convergent evolution of chemical signaling is the best manifested by many similarities between ants and termites, in spite of the great evolutionary distance between these two dominant groups of social insects with great ecological significance (Leonhardt et al., 2016; Tuma et al. 2020).

Since the identification of the multigene family of insect odorant receptors (ORs) expressed in antennae and maxillary palps of Drosophila melanogaster (Clyne et al., 1999), neurophysiology of insect olfaction has seen great progress in terms of phylogenetic reconstructions of OR evolution across taxa, functional characterizations (deorphanisations) of multiple ORs and ultimately their recent structural characterizations (Butterwick et al., 2018; del Mármol et al., 2021; Wang et al., 2024; Zhao et al. 2024). Even though the insect ORs have seven transmembrane domains like the ORs known from vertebrates, they do not share many tangible sequence similarities. Insect ORs have an inverted membrane topology in the dendrites of olfactory sensory neurons compared to vertebrate ORs (Benton et al., 2006; Clyne et al., 1999) and unlike the vertebrate ORs known to act as GPCR receptors, insect ORs function as odorant-gated ion channels (Sato et al., 2008; Wicher et al., 2008). While the ORs from the basal wingless insect lineage Archaeognatha are homotetramers (del Mármol et al., 2021; Brand et al., 2018), ORs of all other Insecta form heteromeric complexes with a highly conserved coreceptor protein (ORCo) (Brand et al., 2018; Butterwick et al., 2018; Larsson et al., 2004; Sato et al., 2008). As recently shown in mosquitoes and aphids, these complexes consist of one OR and three ORCo subunits (Wang et al., 2024; Zhao et al. 2024).

The OR repertoire is greatly variable across Insecta and ranges from units in the wingless Archaeognatha and basal winged order Odonata to tens or hundreds ORs identified in most other flying insects (Robertson, 2019; Yan et al., 2020). Insect ORs often lack a clear orthology pattern across phylogeny, because the OR family evolved via rapid birth-and-death process, accompanied by multiple gene duplications, pseudogenizations, and losses. Lineage-specific expansions, together with a considerable variability in OR ligand specificities (from broad to narrow tuning) allow for a rapid response of olfactory system to ecological and life history changes (Andersson et al., 2015; Benton, 2015; Nei & Rooney, 2005; Robertson, 2019). Ecology-driven plasticity in OR evolution has been convincingly demonstrated by comparisons of specialist vs. generalist insects, the former often having much lower repertoires of ORs and other chemosensory proteins (Robertson, 2019).

The knowledge on OR function and ligand specificities has historically been obtained mainly from deorphanisation studies on Drosophila and other holometabolan insects using various heterologous expression systems. Recently, this bias was in part compensated by OR deorphanisations in more basal taxa, e.g., the wingless Archaeognatha (del Mármol et al., 2021), or the hemimetabolous aphids (Zhang et al., 2017, 2019a) and locusts (Guo et al., 2020; Chang et al., 2023).

Within social insects, eusocial Hymenoptera received considerable attention both in terms of OR repertoire reconstructions and functional characterizations with multiple ORs being deorphanized in ants (Pask et al., 2017; Slone et al., 2017) and the honey bee (e.g., Gomez Ramirez et al., 2023; Wanner et al., 2007). The amassed knowledge suggests that the complex communication and orientation capabilities in the colonies of eusocial Hymenoptera are facilitated by the greatly expanded repertoire of ORs, especially that of the 9-exon subfamily in ants and paper wasps participating in the detection of cuticular hydrocarbons (CHCs) as important cues in contact chemoreception of colony and caste identity and fertility status in eusocial insects (Engsontia et al., 2015; Legan et al., 2021; McKenzie et al., 2016; Pask et al., 2017; Zhou et al., 2015). The 9-exon subfamily and the overall OR richness have been inherited by the extant eusocial Hymenoptera from the ancestor of Aculeata (McKenzie et al., 2016), and also solitary aculeate taxa display large OR repertoires (Obiero et al., 2021); this preadaptation might had been important for the repeated emergence of eusociality and the related complex communication. This is supported by the reduction of OR array (including 9-exon genes) in parasitic ant taxa, along with the simplification of their behavioral repertoire (Jongepier et al., 2022).

The multiple convergences in biology and life histories between termites and ants call for comparison of olfactory detection of chemical signals and environmental cues in the two groups. Yet, despite relatively good knowledge on chemistry of termite pheromones and recognition cues (reviewed in Bagnères & Hanus, 2015; Bordereau & Pasteels, 2011; Mitaka & Akino, 2021), molecular aspects of olfaction remain largely understudied in termites. Termite ORs were so far only addressed with respect to their diversity, inferred from genome assemblies (Harrison et al., 2018; Terrapon et al., 2014) or whole-body transcriptomes (Mitaka et al., 2016), recently complemented by comprehensive search for chemosensory protein repertoire using the antennal transcriptomes of three termite species (Johny et al., 2023). Termite ORs are organized in relatively conserved, highly orthologous pattern and their total numbers range from 28 to 69 (Johny et al., 2023; Terrapon et al., 2014). These numbers are lower than in their solitary cockroach relative Blattella germanica (Harrison et al., 2018), and dramatically lower than in ants, having up to over 400 ORs (Engsontia et al., 2015; Legan et al., 2021; McKenzie et al., 2016; Pask et al., 2017; Zhou et al., 2015). Thus, termites clearly contradict the paradigm on eusociality as a driver of OR richness, even though their chemical communication is by far more complex than in solitary insects and includes pheromone components from a variety of chemical classes, such as fatty-acyl derived alcohols, aldehydes and ketones, terpenoids, and, last, but not least, the CHCs. Independently of eusocial Hymenoptera, termites evolved an intricate communication system using CHCs as kin- and nestmate discrimination cues and as indicators of caste identity and fertility status (reviewed in Bagnères & Hanus, 2015; Mitaka & Akino, 2021). Therefore, compared to ants, the termite social evolution seemingly adopted a different trajectory to accommodate the needs of chemical communication, including CHC detection.

One of alternative hypotheses springs from observations that termites possess an extraordinarily rich set of ionotropic receptors (IRs), reaching up to more than one hundred in some species (Harrison et al., 2018; Johny et al., 2023; Terrapon et al., 2014). Even though the richness of IRs is shared by termites and their cockroach relatives, with B. germanica having the largest IR repertoire ever identified in insects (Robertson et al., 2018), IRs also underwent termite-specific expansions (Harrison et al., 2018; Johny et al., 2023). Since insect IRs have been shown to respond to volatile ligands (Benton et al., 2009), we cannot rule out that they also participate in the complex chemical communication in termites as has been previously speculated (Harrison et al., 2018). Nevertheless, since no chemosensory proteins have yet been functionally characterized in termites, ORs remain the prime candidates for pheromone detection.

Here, we report on the first OR deorphanization in termites. We build on the knowledge about the chemical ecology of the Cuban subterranean termite Prorhinotermes simplex (Rhinotermitidae) (Hanus et al., 2006, 2009; Jirošová et al., 2017; Piskorski et al., 2007), on the annotated repertoire of 50 ORs from P. simplex antennal transcriptome (Johny et al., 2023), on additional P. simplex sequencing data (caste-specific head transcriptomes, draft genome assembly), and on laboratory culture of the species. We select four P. simplex OR sequences, study their function by means of the Empty Neuron Drosophila expression system and single-sensillum recording (SSR) using panels of biologically relevant ligands including the P. simplex pheromones and chemically related compounds. We identify PsimOR14 as the pheromone receptor narrowly tuned to the monocyclic diterpene neocembrene, known as the major component of the trail-following pheromone (TFP) (Sillam-Dussès et al., 2009). We demonstrate strong and selective response of PsimOR14 to neocembrene, with only one additional ligand, geranylgeraniol, generating non-negligible receptor response. We further report on homology-based modelling of neocembrene binding by PsimOR14 and perform molecular dynamics simulations to estimate the impact of ligand binding on PsimOR14 dynamicity. Finally, we identify the P. simplex olfactory sensillum specifically responding to neocembrene and document worker caste-biased expression of PsimOR14 accompanied by significantly higher sensitivity to neocembrene compared to P. simplex soldiers.

Results

Phylogenetic reconstruction and candidate OR selection

In the first step, we reconstructed the phylogeny of termite ORs and ORCos using published protein sequences from two species in combination with our antennal transcriptome data on three species and the bristletail Lepisma saccharina as basal insect outgroup. In the resulting tree, all ORCo sequences and ORs from L. saccharina were basally situated, while majority of termite OR sequences were organized into two large sister clusters, both of which were further split into several sub-branches mostly containing one sequences from all five termite species (Fig. 1A and Supplementary Fig. S1). Only a few exceptions to this highly orthologous pattern were spotted, such as isolated sequences or species-specific expansions with a maximum of four paralogs.

Phylogenetic reconstruction of termite ORs and their transcript abundances in P. simplex workers.

A. Phylogenetic tree is based on 182 protein sequences from five species of termites and the bristletail Lepisma saccharina as a basal insect outgroup, and also includes the sequences of ORCo. The topology and branching supports were inferred using the IQ-TREE maximum likelihood algorithm with the JTT+F+R8 model and supported by 10,000 iterations of ultrafast bootstrap approximation. Protein sequences of termite ORs can be found under the same labeling in Johny et al. (2023). Lepisma saccharina sequences are listed in Thoma et al. (2019). Arrowheads highlight the four ORs from Prorhinotermes simplex selected for functional characterization. Fully annotated version of the tree is provided as Supplementary Fig. S1. B. Heatmap shows the transcript abundances of 50 ORs identified in the RNAseq data from P. simplex worker antennae available in NCBI SRA archive under accession SRX17749141.

Out of the 50 ORs identified in P. simplex, 26 sequences represented full open reading frames with at least 6 undisputed transmembrane domains predicted using TMHMM-2.0. From these, we selected four sequences (PsimOR9, 14, 30 and 31) situated in different parts (subbranches) of the tree (Fig. 1) and used them for transgenic D. melanogaster generation and SSR screening.

Functional characterization of P. simplex ORs in D. melanogaster ab3 sensillum

We expressed the four selected P. simplex ORs in the recently improved version of the Drosophila melanogaster Empty Neuron system (Chahda et al., 2019); the crossing scheme for fly generation adapted from Gonzales et al. (Gonzalez et al., 2016) is shown in Supplementary Fig. S2. Spontaneous SSR firing rates of the four transgenic lines showed an expected pattern with no abnormal bursts, indicating that the ORs were functional. The flies were first subjected to SSR screening with Panel 1, consisting of 11 semiochemicals relevant to termite chemical communication and structurally related compounds. As shown in Fig. 2A, PsimOR9 and PsimOR30 did not provide any strong response to any of the tested compounds, and PsimOR31 broadly and weakly responded to several compounds. By contrast, PsimOR14 systematically and strongly responded to stimulations by the monocyclic diterpene hydrocarbon neocembrene, which is the main component of the trail-following pheromone in the genus Prorhinotermes (Sillam-Dussès et al., 2009). Additionally, a moderate PsimOR14 response was also recorded for the linear diterpene alcohol geranylgeraniol, while all other compounds in the panel, including two other terpenoids, only elicited weak or no responses (Fig. 2A, Supplementary Tables S1–S4).

SSR responses of transgenic D. melanogaster ab3 sensillum expressing PsimOR9, 14, 30 and 31 to the initial screening of 11 volatiles with biological relevance for termites.

A. Heatmap showing the average responses of the four ORs as Δ spikes/s from 3–6 independent replicates. B. Comparison of SSR responses of transgenic Drosophila melanogaster ab3A neurons expressing PsimOR14 (ab3A:PsimOR14) and W1118 D. melanogaster. The bars show the average Δ spikes/s values from five independent replicates ± SEM. C. Characteristic SSR traces of ab3A:PsimOR14 and W1118 flies for 1 µg dose of neocembrene and geranylgeraniol. D. Dose response curve of ab3A:PsimOR14 SSR responses to neocembrene. The graph shows average Δ spikes/s values ± SEM based on nine replicates (8 in case of 100 ng and 4 in case of 500 ng stimulations). The curve fit and ED50 value were calculated using log(agonist) vs. response non-linear algorithm with least square fit method and the constraint of minimal response > 0. The raw data for all graphs is provided in Supplementary Tables S1–S6.

We then compared the responses of Drosophila ab3 sensillum in PsimOR14 expressing flies with those of W1118 flies. As evidenced in Fig. 2B, W1118 ab3 sensillum did not show any significant neuronal response to Panel 1 compounds, while the transgenic PsimOR14 line generated an average Δ spike number of >50 spikes/s for neocembrene and a minor secondary response of ∼25 Δ spikes/s for geranylgeraniol. Characteristic responses for both lines to the two compounds are depicted in Fig. 2C. In the next step, we tested the dose-response behavior of PsimOR14 flies to neocembrene, and recorded an exponentially increasing neuronal response over the range of 0.01–10 ng to ED50 = ∼24 ng and a lack of saturation at the dose of 500 ng (Fig. 2D, Supplementary Tables S5, S6).

PsimOR14 is narrowly tuned to neocembrene, the main trail-following pheromone component

To further address the specificity of PsimOR14 tuning, we tested three additional panels containing 56 frequently occurring insect semiochemicals from various chemical classes. As shown in Fig. 3A, none of these compounds, including multiple terpenoids (mono-, sesqui-, di-), generated a strong response, suggesting a narrow tuning of PsimOR14 to neocembrene. The narrow tuning of PsimOR14 is evident also from the tuning curve depicted in Fig. 3B, with the receptor lifetime sparseness value 0.88 (Supplementary Tables S7–S9). These results confirm that PsimOR14 is a pheromone receptor adaptively tuned to detect the main trail-following pheromone component neocembrene.

SSR responses of transgenic D. melanogaster ab3 sensillum expressing PsimOR14 to the complete set of 67 compounds (Panels 1–4).

A. Heatmap showing the average responses as Δ spikes/s from 3–6 independent replicates. B. Tuning curve of PsimOR14 for the 67 compounds contained in panels 1–4. The raw data is provided in Supplementary Tables S7–S9.

Identification of P. simplex olfactory sensillum responding to neocembrene

Our next goal was to identify the antennal olfactory sensillum responsible for neocembrene detection by P. simplex workers using a combination of SSR measurements with scanning electron microscopy (SEM) and high-resolution SEM (HR-SEM) imaging. Since no previous study reported SSR responses of termite olfactory sensilla to pheromones or environmental cues, we decided to search for the neocembrene-detecting sensillum on the last flagellomere, known to harbor by far the most sensilla in termite workers (Castillo et al., 2021).

In SSR experiments with the termite-relevant compounds from Panel 1, we obtained strong response to both neocembrene and geranylgeraniol from a short multiporous grooved sensillum situated in the apical part of the last antennal segment (Fig. 4A,B). The response of this sensillum to neocembrene was confirmed on workers originating from two different colonies. Detailed view on spontaneous firing pattern of the neocembrene-responding sensillum revealed three different spike amplitudes (Fig. 4C), suggesting the potential presence of as many as three olfactory sensory neurons (a–c). Comparison with spike amplitudes upon neocembrene and geranylgeraniol stimulations then indicated that the responses are generated by the neuron labeled as b.

Neocembrene-responding sensillum in P. simplex workers.

A. SEM photograph of the last flagellomere of P. simplex worker. Arrow shows a small multiporous grooved sensillum responding to neocembrene and geranylgeraniol. Scale bar represents 50 µm. B. HR-SEM view on the neocembrene-responding sensillum. Scale bar in the inset represents 500 nm. C. Detailed view on SSR traces recorded from the neocembrene-responding sensillum during spontaneous firing, and upon stimulation with neocembrene and geranylgeraniol.

The SSR response spectrum of neocembrene sensillum to Panel 1 was markedly similar to that of ab3A neuron of PsimOR14-expressing Drosophila (Fig. 5A). None of the compounds elicited higher average responses than 10 Δ spikes/s, except for neocembrene and geranylgeraniol; their average Δ spikes/s were even slightly higher than those of heterologously expressed PsimOR14, reaching ∼65 and ∼29, respectively (Fig. 5A, B). Likewise, the dose-response experiment with neocembrene indicated a higher sensitivity threshold and lower ED50 = 0.016 ng (Fig. 5C, Supplementary Tables S10, S11).

SSR responses of the neocembrene-responding sensillum on the last flagellomere of P. simplex worker.

A. SSR responses to panel 1. The bars show the average Δ spikes/s values from 8–17 replicates ± SEM. The raw data is provided in Supplementary Table S10. B. Characteristic SSR traces of the neocembrene-detecting sensillum for neocembrene and geranylgeraniol. C. Dose response curve of the SSR responses to neocembrene by the neocembrene-responding sensillum. The graph shows average Δ spikes/s values ± SEM based on 9–11 replicates. The curve fit and ED50 value were calculated using log(agonist) vs. response non-linear algorithm with least square fit method and the constraint of minimal response > 0. The raw data is provided in Supplementary Table S11.

PsimOR14 gene and protein structure, protein modeling, ligand docking, MM/PBSA, molecular dynamics (MD) simulations

Mapping the PsimOR14 transcript sequence on P. simplex draft genome revealed that the gene consists of six exons and is situated on the same locus and in close vicinity of PsimOR15, with which it shares the exon-intron boundaries, suggesting a recent diversification of the two genes via duplication, as also supported by their high sequence similarity (Fig. 6A, Fig. 1A). Transcript (Fig. 6B) and protein (Fig. 6C) structures of PsimOR14 showed the presence of seven transmembrane domains (S1-S7) with the largest extracellular loop between S3 and S4 and the longest intracellular loop between S4 and S5.

PsimOR14 gene, transcript and protein structures, docking and MD simulations.

A. Genomic locus containing PsimOR14 and PsimOR15. PsimOR14 gene consists of 1 non-coding and 5 protein coding exons. B. PsimOR14 transcript with 6 exons, showing the protein-coding (higher boxes) and untranslated regions (lower boxes), and ORF (arrow). C. Transmembrane architecture of PsimOR14. In red are shown seven residues interacting with neocembrene. Light blue ellipse shows the intracellular loop the most impacted by ligand binding. D. Modelled apoform of PsimOR14. Red region denotes the binding site identified via docking, light blue region represents the intracellular S4-S5 loop. E. Holoforms of PsimOR14 with three docked ligands. F. Absolute PsimOR4 dynamicity expressed as average volumes explored by atoms per simulation step in PsimOR14 apoform and upon binding the three studied ligands. G. Relative PsimOR14 dynamicity expressed as average explored atom volumes upon ligand binding relative to the volumes in PsimOR14 apoform. Nucleotide and protein sequences of PsimOR14 are provided in Supplementary Table S12 and as NCBI entry under accession OR921181.

Fig. 6D shows the initial PsimOR14 model obtained using AlphaFold 2. Three terpenoid ligands, i.e., the best agonists neocembrene and geranylgeraniol, and a weak agonist (+)-limonene, were selected for docking into the identified binding site. Dockings are visualized in Fig. 6E, the final scores of the best-ranked poses are shown in Table 1. The predicted docking scores indicated neocembrene as the best ligand, followed by geranylgeraniol and (+)-limonene, in line with the ranking of their biological effect in SSR assays (Table 1). (+)-Limonene ranked as the worst agonist also according to the binding free energy calculated in MM/PBSA analysis, while the best energy score was obtained for geranylgeraniol followed by neocembrene (Table 1). Both docking and MM/PBSA analysis suggested that primarily Van der Waals interactions facilitate the binding, only in case of geranylgeraniol non-negligible contribution of electrostatic interactions has been recorded. Per-residue decomposition results (Fig. 6C, Supplementary Tables S13) showed that all three ligands bind two hydrophobic patches made out of residues from S3 and S4 (Cys154, Ala157, Val158; Thr221, Leu224, Ala225, Tyr228). Neocembrene and (+)-limonene only bind these patches, while geranylgeraniol also interacts with additional residues (Table S13).

MD simulations of per atom explored volumes in PsimOR14 apoform and upon ligand binding delimited three regions with high dynamicity, i.e. the N-terminal region, the extracellular S3-S4 loop and especially the intracellular S4-S5 loop (Fig. 6F). Binding of each of the three ligands reduced the overall protein dynamicity; this protein stabilization did not differ dramatically among the three ligands. By contrast, binding of geranylgeraniol and neocembrene led to a conspicuous dynamicity increase in a portion of the S4-S5 loop, compared to both the apoform and (+)-limonene binding (Fig. 6G).

Caste-biased PsimOR14 expression and antennal sensitivity to neocembrene

We next decided to compare the expression pattern of PsimOR14 between P. simplex workers and soldiers, along with the sensitivity of the two castes to its preferred ligand, neocembrene. Both DESeq2 and EdgeR differential expression analyses of RNAseq read counts from heads (including antennae) of workers and soldiers revealed that PsimOR14 is significantly more expressed in workers, being among three the most upregulated ORs in workers (Fig. 7A, Supplementary Table S14). Subsequent EAG measurements were in line with this observation and indicated significantly stronger responses to neocembrene in workers (p=0.012) (Fig. 7B, Supplementary Table S15).

Caste comparison of PsimOR14 expression and EAG responses between P. simplex workers and soldiers.

A. Volcano plot representing edgeR differential gene expression analysis of all 50 P. simplex ORs in RNAseq data from soldier and workers heads (including antennae) sequenced in three independent biological replicates per caste. Colored dots mark ORs reaching absolute value of log2 fold change ≥ 1, horizontal lines represent p-value thresholds of 0.05 and 0.01. Numeric values of the edgeR and DESeq2 differential expression analysis are provided in Supplementary Table S14. Based on SRA archives accessible under SRX18952230–32 and SRX18952237–39. B. EAG responses of whole antenna preparations of workers and soldiers to neocembrene at a dose of 10 ng (mean ± SD shown on log2 scale). Inter-caste differences were compared using t-test on log2-transformed data. Raw data is shown in Supplementary Table S15.

Discussion

Identification of PsimOR14 as the pheromone receptor narrowly tuned to neocembrene in P. simplex represents the first OR deorphanisation in termites and confirms that the trail-following communication is mediated by odorant receptors. This assumption, validated here for the monocyclic diterpene neocembrene, is indirectly supported by previous experiments in two termite species having a C12 fatty alcohol as TFP; ORCo silencing in these species impaired the ability to follow the foraging trail (Gao et al., 2020). Because trail-following in termites has the shared evolutionary origin with courtship communication and both neocembrene and C12 alcohols also occur as sex-pairing pheromone components (Bagnères & Hanus, 2015; Bordereau & Pasteels, 2011; Sillam-Dussès, 2010), it is likely that the two communication modalities share identical or closely related ORs. Future OR deorphanizations should test this hypothesis and provide an insight on how the functional diversification of trail-following and sex attraction is imprinted into the OR evolution.

Neurophysiological characteristics of D. melanogaster ab3A neuron expressing PsimOR14 showed expected patterns of spontaneous firing rates of units of Hz (6.87 ± 4.73, mean ± SD) and maximum firing rates of 90 spikes/s at the highest used neocembrene doses (1 µg), though not reaching the reported maxima for ab3A responses with genuine and exogenous receptors, which may be well over 100 spikes/s (e.g., Chahda et al., 2019). This confirms that co-expression of termite ORs with the Drosophila ORCo using the Empty Neuron system is a suitable technique for termite OR characterizations. PsimOR14 has a narrow tuning to neocembrene with receptor lifetime sparseness equal to 0.88. ORs detecting pheromone components (pheromone receptors) and other volatiles with high biological importance, such as key food or host attractants, are often expressed in specialized sensilla and are highly selective, in contrast to the broad tuning of ORs sensing the general environmental stimuli (e.g., Carey et al., 2010; Fleischer & Krieger, 2018). Such a high selectivity is the best known for sex pheromone receptors, e.g. in moths (reviewed in Zhang & Löfstedt, 2015) or Drosophila (reviewed in Haverkamp, et al. 2018), mostly tuned to a single compound, even though in some cases the respective ORs can be adaptively shaped to selectively detect more pheromone compounds (Díaz-Morales et al., 2024; Mariette et al., 2024). Thus, the narrow tuning of PsimOR14 is in line with expectations from a pheromone receptor, especially when considering the ancient origin of trail-following from sex-pairing behavior and the great importance of TFP for concerted foraging in these blind insects. Nevertheless, due to the low coverage of insect diversity by OR deorphanization studies and their strong bias towards derived Holometabola, it remains difficult to make any general considerations about the OR selectivity relative to the receptor function across Insecta. This has been recently demonstrated by a comprehensive functional characterization of ORs in migratory locust unveiling an unexpected design of the olfactory information decoding, consisting of a large set of narrowly tuned ORs for environmental cues (Chang et al., 2023).

With average Δ spike/s over 20, the linear diterpene alcohol geranylgeraniol was the only other agonist with non-negligible response. None of the remaining compounds elicited a Δ spike of more than 7, despite the presence of multiple other compounds derived from terpenoid scaffolds in the tested panels. It is difficult to attribute any adaptive significance to this observation since there is no record of geranylgeraniol in communication context in termites. Therefore, the observed significant biological activity, high docking score and high ranking in free energy comparison, may be due to non-adaptive binding affinity of PsimOR14 to the non-native ligand (see also below). Interestingly, another related terpene alcohol, the monoterpene linalool, had the biggest negative Δ spike score (-6.8) and the absolute spikes/s ranged from 0 to 5 (1.5 ± 1.9, mean ± SD), suggesting a possible inverse agonist function of linalool.

The neocembrene-sensing sensilla identified using SSR in P. simplex workers provided a response pattern to Panel 1 very similar to that of PsimOR14-expressing ab3 Drosophila sensillum, e.g., strong response to neocembrene followed by geranylgeraniol, supporting independently the results obtained from transgenic Drosophila. No other important responses to termite pheromone components from the Panel 1, which also included the minor P. simplex TFP component (3Z,6Z,8E)-dodecatrien-1-ol, were recorded, in spite of the likely presence of two additional olfactory neurons housed in the neocembrene-detecting sensillum. The function of these additional neurons remains elusive and examples from other insects offer multiple scenarios. A neuron expressing a highly selective pheromone receptor may co-exist with functionally independent neurons detecting environmental cues (e.g., Tateishi et al. 2020), or the co-habitation of several neurons may generate more complex interactions such as lateral inhibition of the neighboring neurons (Su et al., 2012; Zhang et al., 2019b). However, it must be noted, that our SSR analyses focused a specific sensillum defined by its topology on the last flagellomere and we did not perform a comprehensive mapping of olfactory sensilla on the entire worker antenna. Therefore, also our image on the distribution of neocembrene-detecting sensilla and the pattern of neurons housed in these sensilla is incomplete.

The basic PsimOR14 protein architecture with seven transmembrane domains is typical for insect ORs. Likewise, the docking experiments identified a binding site defined by a binding pocket deep in the transmembrane region, homologous to that in previously studied insect ORs, and confirmed the nature of the ligand binding to mainly rely on hydrophobic interactions (e.g., del Mármol et al., 2021; Pettersson & Cattaneo, 2023; Yuvaraj et al., 2021). In case of PsimOR14, mainly the residues from S3 and S4 interacted with the studied ligands; these transmembrane domains were shown to participate in ligand binding also in other studied insect ORs (e.g., del Mármol et al., 2021; Yuvaraj et al., 2021; Wang et al., 2024; Zhao et al. 2024).

The calculated binding scores for the three tested ligands correlated with observed biological effects of the ligands. The MM/PBSA data also in part corroborated the SSR measurements by estimating (+)-limonene as the worst agonist; by contrast, geranylgeraniol and not neocembrene ranked as the best ligand. This may potentially be explained by the additional residues interacting with geranylgeraniol compared to neocembrene and (+)-limonene, and the relatively higher contribution of electrostatic interactions in geranylgeraniol binding. Moreover, in systems where ligand binding induces allosteric effects, as is the case of insect ORs, the simple binding affinity does not provide a complete picture of receptor signal transduction function.

In contrast to the general protein stabilization upon binding of the three tested ligands, some parts of the intracellular S4-S5 loop showed a significant increase in the dynamics upon binding of the biologically active ligands geranylgeraniol and neocembrene, compared to apoform and the weak agonist (+)-limonene binding. What is the mechanistic role of this allosteric transmission of the binding effect to the intracellular loop remains elusive. However, the impact of the S4-S5 loop on OR function has previously been reported for the MhOR5 from Machilis hrabei (del Mármol et al., 2021); replacement of the loop with a short linker increased the receptor response to native ligand. Interestingly, in some insect ORs this loop has been reduced to a short sequence of a few residues (e.g., Wang et al., 2024).

PsimOR14 does not belong to the most expressed P. simplex ORs (Fig. 1B). However, interestingly, it is among those having the most caste-biased expression with significantly higher transcript abundance in antennae of workers compared to soldiers. Accordingly, also the electrophysiological responses of workers to neocembrene were significantly stronger in workers.

Termite soldiers are known to lay pheromone trails, to detect TFPs and to participate in foraging and field exploration in a number of termite species, though the behavioral patterns of soldiers and workers during these activities differ (Kaib, 1990; Traniello, 1981; Traniello & Busher, 1985). This has also been demonstrated in a close relative to our model, the congeneric species Prorhinotermes inopinatus (Rupf & Roisin, 2008). Yet, differences in sensitivity of termite workers and soldiers have not previously been addressed at the electrophysiological level. Caste-biased PsimOR14 expression and neocembrene sensitivity may represent olfactory information filtering adaptive to the different tasks of the two castes, as documented, e.g., in ants (Caminer et al., 2023).

Genus Prorhinotermes is the most basally situated termite taxon known to have a terpenoid component as a part of its pheromone repertoire (Sillam-Dussès et al., 2009). The acquisition of terpenoid pheromone components is undoubtedly due to the evolution of terpene biosynthesis in basal Neoisoptera, which has led to the fascinating diversity of defensive terpenoids produced by soldiers of Neoisoptera in their frontal gland (Gössinger, 2019). In line with general observations on pheromone evolution (Steiger et al., 2010), some of these defensive terpenoids from the frontal gland gained via exaptation a novel function of alarm pheromones (Dolejšová et al., 2014; Roisin et al., 1990; Šobotník et al., 2008, 2010). Likewise, the cyclic diterpenes probably became part of trail-following pheromones and sex-pairing pheromones by co-opting the terpenoid biosynthesis in exocrine glands other than the defensive frontal gland of soldiers. Beside the occurrence of neocembrene in the sub-basal Prorhinotermes, it only occurs as a pheromone component in much later diverging representatives of the subfamily Nasutitermitinae (Bordereau & Pasteels, 2011). Once reliable genome or antennal transcriptome of Nasutitermitinae become available, it would be of interest to search in these species for OR orthologs of PsimOR14 and test whether they have retained the function of neocembrene detection.

Insect ORs are frequently organized in tandem arrays (e.g., Bohbot et al., 2007; McKenzie and Kronauer, 2018; Robertson and Wanner, 2006). Likewise, the genomic locus of PsimOR14 contains a likely tandem copy paralog of PsimOR15 gene. The two genes share the gene architecture, and their transcripts are equally represented in worker antennal transcriptome, though PsimOR15 does not display the caste-biased expression. Thanks to the close similarity of both genes, PsimOR15 is a candidate for potential pheromone receptor function.

Future research in termites should also aim at finding the ORs or other chemosensory proteins involved in the detection of CHCs. In spite of the similarities in CHC roles in termites and ants, ranging from nestmate recognition to fertility signaling, termite OR sequences do not show any conspicuous expansions analogous to 9-exon subfamily in ants. Therefore, finding the chemosensory principle of CHC detection in termites would bring another piece of knowledge about the fascinating convergent evolution in these two unrelated major groups of eusocial insects. Yet another appealing target for OR deorphanization is the search for pheromone receptors of queen pheromones, the central signals ensuring the reproductive dominance of queens in the colonies of social insects. In the honeybee and ants, narrowly tuned ORs responding to main queen pheromone components were already described (Wanner et al., 2007; Pask et al., 2017). In termites, queen pheromones were so far only identified in two species (Dolejšová et al., 2022; Matsuura et al. 2010). They are volatile and were shown to act as airborne signals via olfaction (Dolejšová et al., 2022), it is thus reasonable to expect a selective OR responsible for their detection.

Materials and methods

Termites

Multiple laboratory colonies of P. simplex, originating from previous field collections in Cuba and Florida, are held in the Institute of Organic Chemistry and Biochemistry, Czech Academy of Sciences. Colonies are reared in glass vivaria at 27°C and 80% relative humidity in clusters of spruce wood slices.

The data reported here were collected from three mature Cuban colonies. The first one was used for antennal transcriptome sequencing and assembly followed by phylogenetic analysis, as described in Johny et al. (2023). The second one was used for RNA extraction for OR cloning, SSR, SEM and HR-SEM. The third one was used for caste-specific head transcriptomes (head + antennae) of workers and soldiers, for caste-specific EAG recordings, and for SSR confirmation of the neocembrene-detecting sensillum. For all experiments with workers, 4th or 5th stage workers were selected as the most abundant developmental stages, recognized according to body size and head width.

RNA extraction, OR cloning and construct generation

Total RNA was extracted from 20 pairs of dissected P. simplex antennae using PureLink RNA Mini Kit (Invitrogen, Carlsbad, CA, USA) following the manufacturer’s protocol and quantified using NanoDrop spectrophotometer (Thermo, Delaware, USA). From the total RNA, 2 µg was used to synthesize the cDNA using SuperScript IV Reverse Transcriptase (Invitrogen, Carlsbad, CA, USA) according to manufacturer’s instructions. The efficiency of cDNA synthesis was evaluated by amplification of ORCo. The list of primers is provided in Supplementary Table S16. The full-length open reading frame (ORF) of each selected PsimORs was PCR-amplified from the cDNA using the DreamTaq Green PCR Master Mix (Invitrogen, USA) and gene-specific primers (Supplementary Table S16). Amplification products were purified by QIAquick Gel Extraction Kit (Qiagen, Germany), cloned into pCR8/GW/TOPO vector using the TOPO TA Cloning Kit (Invitrogen, USA) and transformed into OneShot TOP10 competent cells (Invitrogen, USA). Positive colonies were selected based on colony-PCRs using primers GW1 and GW2, recombinant plasmids were isolated using the QIAprep 2.0 Spin Miniprep Columns (Qiagen, Germany) and sequences were verified by Sanger sequencing (Eurofins Genomics, Germany).

The expression vector constructs were prepared using the Gateway LR recombination cloning technology (Invitrogen, USA) based on recombination of the phage-like attachment sites attL/R in pCR8/GW/TOPO with the bacteria-like attachment site attB in pUASg.attb vector (obtained from Drosophila Genomics Resource Center, Bloomington, USA). The resulting constructs pUASg.attB-PsimOR were purified using the QIAprep 2.0 Spin Miniprep Columns (Qiagen, Germany) and insert sequences were verified by Sanger sequencing at Eurofins Genomics (Germany). All primers used for Sanger sequencing and colony-PCR are listed in Supplementary Table S16).

Fly lines

D. melanogaster lines used in the Empty Neuron system were kindly provided by Dr. Thomas O. Auer (from Richard Benton Lab, University of Lausanne, Switzerland). The wild type W1118 line, used as a control, was kindly provided by Prof. Michal Žurovec (Biology Centre, Czechia). All Drosophila melanogaster lines were reared in an incubator which was set at 24±2°C with relative humidity of 50±5%. Flies were fed with in-house prepared diet based on standard cornmeal food. The fly lines used are listed in Supplementary Table S17.

Transgenic expression of termite ORs in D. melanogaster ab3A neuron

Selected PsimORs were expressed in the Drosophila melanogaster Empty Neuron system for functional screening. Transgenic D. melanogaster UAS-PsimOR lines were generated by BestGene Inc. (Chino Hills, CA, USA) by injecting pUASg.attB-PsimOR vectors into fly embryos expressing the integrase PhiC31 and carrying an attP landing site resulting in flies with genotype w−; +; UAS-PsimOR (w+)/+.

The recent CRISPR-Cas9-engineered empty neuron line Or22ab-Gal4 (Chahda et al., 2019) was used as Δhalo genetic background for the expression of UAS-PsimOR in Dmel ab3 sensilla. The fly crossing scheme was adapted from (Gonzalez et al., 2016) with a modification at the F3 crossing. Final homozygote lines with UAS-PsimOR and Or22ab-Gal4 were generated and used for the electrophysiological recordings. The full description of the crossing scheme is provided in Supplementary Fig. S2.

Chemicals

For SSR measurements, we used a total of 67 chemicals organized into four panels. The initial screening Panel 1 contained 11 compounds biologically relevant to termites, i.e. components of termite pheromones and cuticular hydrocarbons known from Neoisoptera and structurally related compounds. This panel was used for initial SSR screening of PsimOR9, 14, 30 and 31 in transgenic D. melanogaster and for SSR experiments with P. simplex workers. For detailed analysis of Psim OR14, three additional panels were used, consisting of 56 frequently occurring insect semiochemicals (e.g., terpenoids, fatty acid esters, fatty alcohols and aldehydes, etc.). Panel 1 compounds were diluted in n-hexane to 100 ng/µL, panel 2–4 compounds were diluted in paraffin oil to 10-3 v/v. List of all compounds tested and their origin are listed in Supplementary Table S18.

Organic synthesis

For the purpose of SSR experiments, we synthesized (Z)-dodec-3-en-1-ol, (3Z,6Z)-dodeca-3,6-dien-1-ol, (3Z,6Z,8E)-dodecatrien-1-ol, and dodec-3-yn-1-ol, and included these compounds into Panel 1. The de novo organic synthesis of these fatty alcohols is described in the Supplementary Information file.

Electrophysiology

SSR recordings on Drosophila ab3 sensillum were performed as described previously (Benton & Dahanukar, 2023; Olsson & Hansson, 2013). We used 2–4 days old flies for one recording per each to avoid neuronal adaptations from multiple stimulations. To expose more ab3 sensilla, the fly preparation was done with arista down (Keesey et al., 2022).

In termites, the olfactory sensilla situated on the last antennal flagellomere of workers were targeted for SSR, since their number increases towards the distal end of termite antennae, the last segment being significantly more populated by olfactory sensilla than any other segment (Castillo et al., 2021). The grounding electrode was carefully inserted into the clypeus and the antenna was fixed on a microscope slide using a glass electrode. To avoid the antennal movement, the microscope slide was covered with double-sided tape and the three distal antennal segments were attached to the slide.

The sensilla were observed under the Nikon FN1 eclipse microscope at 60× magnification. For all electrophysiological measurements, the recording electrode was brought into contact with the base of the sensillum using a Kleindiek Nanotechnik MM3A micromanipulator connected to a cubic micromanipulator device. Using Syntech stimulus delivery system (CS55 model, Syntech, Germany), the odorant stimulus was administered as a 0.3 s pulse by inserting the tip of the glass Pasteur pipette through an opening into the delivery tube (situated 4 cm from the tube outlet) carrying a purified air stream (0.4 L/min). The tube outlet was placed approximately 1 cm from the antenna. The experiments were conducted at 25–26 °C.

From each diluted odorant (100 ng/µL), 10 µL were pipetted on 1 cm diameter filter paper disk placed in a glass Pasteur pipette in the screening experiment, while the doses ranging from 0.01 to 500 ng of neocembrene per filter paper were used in the dose-response experiments.

The signal was amplified and digitally converted using Syntech IDAC-4. The neuronal cells were sorted based on their amplitude and the spikes were counted using the AutoSpike v3.9software (Syntech Ockenfels, Germany). Δ spike was calculated by subtracting the number of spikes during 1 second post-stimulation from the number of spikes generated 1 second before the stimulation. In dose response experiments and measurements on termite antennae, the counting periods were 0.5s. Δ spike values were corrected by subtracting the response generated by the solvent and converted into Δ spike/s (Benton & Dahanukar, 2023; Olsson & Hansson, 2013). The receptor lifetime sparseness value was calculated according to Chang et al. (2023).

EAG experiment addressing the caste-specificity of antennal responses to neocembrene was performed with 15 workers and 15 soldiers; each individual was only used for one stimulation series consisting of air–hexane–neocembrene (10 ng)–hexane–air. Brain and antennal tip were placed between two Ag/AgCl electrodes containing Ringer’s solution and connected to a high impedance (1014 Ω) amplifier (Syntech, Buchenbach, Germany). The antennal preparation was placed into a stream of cleaned air (500 mL/min), into which the stimuli were injected from Pasteur pipettes containing a 1.5 cm2 filter paper impregnated with 10 µL of the tested solution. Odor injections were controlled by a foot switch-operated Syntech stimulus controller and maximal negative deflection was measured using Syntech EagPro software. Pasteur pipettes containing odorant stimuli were changed after three stimulations. Air responses were used for data normalization, the responses log2-transformed to reduce heteroscedasticity and comply with assumptions for parametric testing (Bartlett test for equal variances, and Shapiro-Wilk normality test), and then compared between workers and soldiers using Student’s t-test.

Scanning electron microscopy

For SEM, 10 workers with intact antennae were cold-anesthetized and decapitated with micro-scissors. Heads were desiccated in increasing ethanol concentrations (60, 80, 90, and 96%, each for 2 h) followed by 12 h in acetone. Heads were then attached to aluminium holders for microscopy using adhesive tape, and differentially oriented to allow axial, dorsal, ventral, and lateral views. The samples were gold-coated for regular SEM (4 nm gold layer) and high-resolution SEM (HR-SEM, 2 nm) using sputter coater Bal-Tec SCD 050. Last antennal segments were inspected and photographed under scanning electron microscope JEOL 6380 LV (SEM). Surface of particular sensilla were studied using high resolution field emission scanning electron microscope JSM-IT800 (HR-SEM) and Olympus Soft Imaging Solution software. Working distance for all samples was 4.0–4.1 mm and accelerating voltage 2.0 kV.

Bioinformatics

For phylogenetic reconstruction of termite ORs, we used 182 OR protein sequences originating from five termite species, i.e. Neotermes cubanus, Prorhinotermes simplex, Inquilinitermes inquilinus (Johny et al., 2023), Zootermopsis nevadensis (Terrapon et al., 2014) and Cryptotermes secundus (Harrison et al., 2018), and the bristletail Lepisma saccharina (Thoma et al., 2019) as a basal insect outgroup. For all species also the ORCo sequence was included. Sequences were aligned by means of the MUSCLE algorithm and used for reconstructing the phylogenetic tree with the IQ-TREE maximum likelihood algorithm (Nguyen et al., 2015) using the JTT+F+R8 substitution model and 10,000 ultrafast bootstrap replicates.

The gene structures of PsimORs were characterized by local alignment of full-length transcript sequences from Johny et al. (2023) to our in-house genome assembly using BLAST (for details on genome assembly see Koubová et al., 2021) and confirmed with genomic mapping of the RNAseq data from P. simplex antennae available under accession SRX17749141 in NCBI SRA archives using STAR aligner v2.7.10b (Dobin et al., 2013). The mapping results were further used for abundance estimations of all ORs and ORCo in antennal transcriptome reported in Johny et al. (2023). Read counts were obtained with featureCounts tool from the Subread package (https://subread.sourceforge.net/) and normalized according to the FPKM (Fragments Per Kilobase Million) method.

Differential OR expression analysis in P. simplex soldier and worker heads (including antennae) was performed using the RNAseq data from our previous sequencing project and available as SRA archives under accessions SRX18952230–32 and SRX18952237–39. The data was obtained from sequencing of three biological replicates of each caste. Read counts obtained using STAR mapping and featureCounts estimations were statistically evaluated using the DESeq2 Bioconductor package in R and edgeR.

PsimOR14 secondary structure was predicted using online tools Jpred 4.0.0 (www.compbio.dundee.ac.uk/jpred) and TMHMM2.0 (https://services.healthtech.dtu.dk/services/TMHMM-2.0), schematic model was generated using Protter (https://wlab.ethz.ch/protter).

Protein modeling

PsimOR14 structure was modelled in its monomeric membrane-free form using AlphaFold2 (Jumper et al., 2021; Mirdita et al., 2022). The best model was refined by MD relaxation, employing GROMACS 2021.3 and CHARMM36m (Abraham et al., 2015; Huang et al., 2017). After solvation and neutralization by Na+ ions in TIP3P CHARMM water in a 1.5 nm padded box, temperature and pressure equilibration followed. Six different simulations with differing starting velocities were produced, followed by 150 ns periodic simulated annealing independently for each replica (0.5 ns at 300 K, then 0.5 ns at 320 K, repeating). The lowest potential energy structure was chosen. Neocembrene, geranylgeraniol and (+)-limonene structures were sourced from PubChem (13th Feb 2023) and parametrized using CgenFF 4.6, employing CHARMM-GUI for the conversion (Jo et al., 2008; Kim et al., 2023; Vanommeslaeghe et al., 2010). The binding site was predicted based on DEET binding region of MhOR5 from Machilis hrabei (7LIG) (del Mármol et al., 2021). Using DockThor webserver, ligands were docked with all bonds treated as rotatable, centered around the expected binding site with maximized box size (40 units). The best binder was then selected for each complex (Santos et al., 2020).

MM/PBSA simulations

Complex topologies were built in GROMACS. Simulations included 9 replicas for liganded (14 µs each) and 6 for unliganded PsimOR14 (6.9 µs), with different starting velocities. Convergence was assessed by ligand backbone RMSD distributions. Replicas were concatenated, split into 1.5 ns frames, and analyzed with MM/PBSA (ff19SB+GAFF2, linearized PB with diel=2, SASA for apolar contributions, optimized CHARMM radii) using gmx_MMPBSA v1.6.3 and AmberTools 20 (Case et al., 2023; Guedes et al., 2021; Tian et al., 2020; Wang et al., 2004). For each liganded PsimOR14 (neocembrene, geranylgeraniol and limonene) 9 replica trajectories were separately analyzed, along with apoform PsimOR14 trajectories (6 replicas). All trajectories were PBC-corrected, cleaned, and fitted to the first frame of the PsimOR14 trajectory.

Atom positions were marked in 3D space per frame, and the convex hull algorithm approximated the volume explored by each atom. This compared dynamic behavior between trajectories, with unliganded PsimOR14 as the baseline. The average total volume explored by atoms was compared.